---
title: "C3 Marker Browser Montages for Selected ESC Transition Studies"
subtitle: "Real gene models, BigWig-derived normalization, top up/down C3 peaks, average and replicate-level browser views"
output:
  html_notebook:
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: inline
---

This notebook rebuilds the selected-study C3 browser figures using the BigWigs directly.

Key design choices:

1. Peak ranking is computed directly from normalized BigWig signal over strict `C3` marker peaks.
2. Loci are selected programmatically as top `up` and `down` `C3` peaks for each study comparison.
3. Real gene models are drawn with `Gviz::GeneRegionTrack` from `TxDb.Mmusculus.UCSC.mm10.knownGene`.
4. Main figure uses replicate averages for readability.
5. Companion plots keep all replicates.
6. Two scaling modes are saved:
   - `absolute_normalized`: library-size normalized signal, with panel y-limits set from the non-epiblast tracks to avoid flattening everything else
   - `autoscaled`: each track divided by its own 99th percentile, for shape comparison only

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
renv::activate()

cran_pkgs <- c("tidyverse", "patchwork", "glue", "scales", "png", "svglite", "xml2")
bioc_pkgs <- c(
  "GenomicRanges", "rtracklayer", "Signac", "Gviz",
  "TxDb.Mmusculus.UCSC.mm10.knownGene", "org.Mm.eg.db"
)

for (pkg in cran_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg, ask = FALSE, update = FALSE)
}

suppressMessages({
  library(tidyverse)
  library(patchwork)
  library(glue)
  library(scales)
  library(png)
  library(svglite)
  library(xml2)
  library(grid)
  library(GenomicRanges)
  library(rtracklayer)
  library(Signac)
  library(Gviz)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Mm.eg.db)
})

set.seed(42)
options(ucscChromosomeNames = FALSE)
```

## 2. Configuration

```{r config}
DATA_DIR <- file.path("..", "data")
RESULTS_DIR <- file.path("..", "results")
METADATA_FILE <- file.path("..", "ATAC_bigwig_metadata_AIannotated.csv")
MARKER_FILE <- file.path(RESULTS_DIR, "Epiblast_clusters_marker_peaks_annotated_reg.tsv")

OUT_ROOT <- file.path(RESULTS_DIR, "C3_marker_browser_montage_selected_studies")
OUT_TABLES <- file.path(OUT_ROOT, "tables")
OUT_LOCI <- file.path(OUT_ROOT, "locus_panels")
OUT_FIGS <- file.path(OUT_ROOT, "figures")

for (d in c(OUT_ROOT, OUT_TABLES, OUT_LOCI, OUT_FIGS)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

state_colors <- c("ESC" = "#4D4D4D", "EpiLC" = "#D55E00", "EpiSC" = "#007C91")
cluster_colors <- c("C1" = "#3F84AA", "C2" = "#DABE99", "C3" = "#65A83E", "C4" = "#F397C0")
study_track_colors <- c(
  "E6.5 Epiblast_C3" = unname(cluster_colors["C3"]),
  "ESC" = unname(state_colors["ESC"]),
  "EpiLC" = unname(state_colors["EpiLC"]),
  "EpiSC" = unname(state_colors["EpiSC"]),
  "sEpiLC" = "#F0B35A",
  "d2 EpiLC" = unname(state_colors["EpiLC"]),
  "EpiLC+CHIR" = "#7AAE3A",
  "FTW" = "#B85C8C"
)

marker_log2fc_cutoff <- 0.5
marker_padj_cutoff <- 0.05
top_peaks_per_direction <- 3
all_c3_chunk_size <- 6
selected_gene_chunk_size <- 4
region_flank_factor <- 5
region_min_flank_bp <- 3000
selected_gene_fixed_window_bp <- 80000
bin_size_bp <- 25
selected_gene_bin_size_bp <- 25
selected_gene_smoothing_window_bp <- 1000
selected_gene_png_width_px <- 1080
selected_gene_svg_width_in <- 6.3
peak_highlight_fill <- "#F4D03F35"
stable_log2fc_threshold <- log2(1.2)
selected_gene_priority <- c("Nfatc1", "Aplnr", "Mesp1", "Mixl1", "Snai1", "Cdh4", "Foxj2", "Gata6")
```

## 3. Helper Functions

```{r helpers}
clean_str <- function(x) {
  x <- ifelse(is.na(x), "", x)
  gsub("\\s+", " ", trimws(x))
}

sanitize_numeric_matrix <- function(mat) {
  mat <- as.matrix(mat)
  storage.mode(mat) <- "double"
  mat[!is.finite(mat)] <- NA_real_
  mat
}

compute_size_factors <- function(mat) {
  mat <- sanitize_numeric_matrix(mat)
  sample_medians <- apply(mat, 1, function(x) median(x[is.finite(x)], na.rm = TRUE))
  pos <- sample_medians[is.finite(sample_medians) & sample_medians > 0]
  gm <- median(pos, na.rm = TRUE)
  if (!is.finite(gm) || gm <= 0) gm <- 1
  sf <- sample_medians / gm
  sf[!is.finite(sf) | sf <= 0] <- 1
  names(sf) <- rownames(mat)
  sf
}

extract_primary_gene <- function(x) {
  xs <- strsplit(ifelse(is.na(x), "", x), "[,;|/ ]+")
  vapply(xs, function(y) {
    y <- y[nzchar(y)]
    if (length(y) == 0) NA_character_ else y[1]
  }, character(1))
}

harmonize_gene_symbols <- function(x) {
  x <- ifelse(is.na(x), "", x)
  out <- x

  ens_idx <- grepl("^ENSMUSG", out)
  if (any(ens_idx)) {
    ens_map <- AnnotationDbi::mapIds(
      org.Mm.eg.db,
      keys = unique(out[ens_idx]),
      column = "SYMBOL",
      keytype = "ENSEMBL",
      multiVals = "first"
    )
    out[ens_idx] <- ifelse(
      !is.na(ens_map[out[ens_idx]]) & nzchar(ens_map[out[ens_idx]]),
      unname(ens_map[out[ens_idx]]),
      out[ens_idx]
    )
  }

  entrez_idx <- grepl("^[0-9]+$", out)
  if (any(entrez_idx)) {
    entrez_map <- AnnotationDbi::mapIds(
      org.Mm.eg.db,
      keys = unique(out[entrez_idx]),
      column = "SYMBOL",
      keytype = "ENTREZID",
      multiVals = "first"
    )
    out[entrez_idx] <- ifelse(
      !is.na(entrez_map[out[entrez_idx]]) & nzchar(entrez_map[out[entrez_idx]]),
      unname(entrez_map[out[entrez_idx]]),
      out[entrez_idx]
    )
  }

  out[out == ""] <- NA_character_
  out
}

bw_mean_over_ranges <- function(bw_path, gr) {
  cov <- rtracklayer::import(bw_path, format = "BigWig", which = gr, as = "RleList")
  vals <- rep(NA_real_, length(gr))
  gr_chr <- as.character(seqnames(gr))
  cov_chr <- names(cov)

  mapped <- dplyr::case_when(
    gr_chr %in% cov_chr ~ gr_chr,
    paste0("chr", gr_chr) %in% cov_chr ~ paste0("chr", gr_chr),
    sub("^chr", "", gr_chr) %in% cov_chr ~ sub("^chr", "", gr_chr),
    TRUE ~ NA_character_
  )

  valid_idx <- which(!is.na(mapped))
  if (length(valid_idx) > 0) {
    by_chr <- split(valid_idx, mapped[valid_idx])
    for (chr in names(by_chr)) {
      idx <- by_chr[[chr]]
      v <- Views(cov[[chr]], start = start(gr[idx]), end = end(gr[idx]))
      vals[idx] <- viewMeans(v, na.rm = TRUE)
    }
  }

  vals[is.na(vals)] <- 0
  vals
}

extract_region_signal <- function(bw_path, seq_chr, start_bp, end_bp) {
  gr <- GRanges(seqnames = seq_chr, ranges = IRanges(start = start_bp, end = end_bp))
  cov <- rtracklayer::import(bw_path, format = "BigWig", which = gr, as = "RleList")
  cov_chr <- names(cov)
  mapped_chr <- dplyr::case_when(
    seq_chr %in% cov_chr ~ seq_chr,
    paste0("chr", seq_chr) %in% cov_chr ~ paste0("chr", seq_chr),
    sub("^chr", "", seq_chr) %in% cov_chr ~ sub("^chr", "", seq_chr),
    TRUE ~ NA_character_
  )
  if (is.na(mapped_chr)) return(rep(NA_real_, end_bp - start_bp + 1))
  as.numeric(cov[[mapped_chr]][start_bp:end_bp])
}

bin_signal <- function(signal_vec, chr, from_bp, bin_size = 25) {
  idx <- seq_along(signal_vec)
  bins <- split(idx, ceiling(idx / bin_size))
  tibble(
    seqnames = chr,
    start = from_bp + vapply(bins, min, integer(1)) - 1L,
    end = from_bp + vapply(bins, max, integer(1)) - 1L,
    signal = vapply(bins, function(ii) mean(signal_vec[ii], na.rm = TRUE), numeric(1))
  )
}

smooth_signal <- function(signal_vec, window_bp = 1000) {
  if (is.null(window_bp) || !is.finite(window_bp) || window_bp <= 1) {
    return(signal_vec)
  }

  k <- as.integer(round(window_bp))
  if (k %% 2 == 0) k <- k + 1L
  k <- min(k, length(signal_vec))
  if (k <= 1) return(signal_vec)

  kernel <- rep(1 / k, k)
  smoothed <- as.numeric(stats::filter(signal_vec, kernel, sides = 2))

  if (all(is.na(smoothed))) {
    return(signal_vec)
  }

  valid <- which(!is.na(smoothed))
  if (length(valid) > 0) {
    first_valid <- valid[1]
    last_valid <- valid[length(valid)]
    if (first_valid > 1) smoothed[seq_len(first_valid - 1)] <- smoothed[first_valid]
    if (last_valid < length(smoothed)) smoothed[seq.int(last_valid + 1, length(smoothed))] <- smoothed[last_valid]
  }

  smoothed[is.na(smoothed)] <- signal_vec[is.na(smoothed)]
  smoothed
}

make_image_plot <- function(img_path) {
  patchwork::wrap_elements(
    full = grid::rasterGrob(png::readPNG(img_path), interpolate = TRUE)
  )
}

distinct_primary_gene <- function(df) {
  df %>%
    mutate(primary_gene = ifelse(is.na(primary_gene) | primary_gene == "", region, primary_gene)) %>%
    group_by(primary_gene) %>%
    slice_head(n = 1) %>%
    ungroup()
}

format_pct_delta <- function(x) {
  if (!is.finite(x)) return("NA")
  paste0(ifelse(x > 0, "+", ""), sprintf("%.0f%%", x))
}

simplify_track_label <- function(x) {
  dplyr::case_when(
    x == "E6.5 Epiblast_C3" ~ "C3",
    x == "EpiLC+CHIR" ~ "CHIR",
    TRUE ~ x
  )
}

classify_direction <- function(log2fc, threshold = stable_log2fc_threshold) {
  dplyr::case_when(
    log2fc >= threshold ~ "up",
    log2fc <= -threshold ~ "down",
    TRUE ~ "similar"
  )
}

render_locus_gviz <- function(
  peak_rows,
  track_specs,
  scaling_mode = c("absolute_normalized", "autoscaled"),
  out_png = NULL,
  out_svg = NULL,
  title_prefix,
  compact = FALSE,
  label_vs_epiblast = FALSE,
  fixed_window_bp = NULL,
  bin_size_local_bp = bin_size_bp,
  smoothing_window_bp = NULL,
  png_width_px = 1800,
  svg_width_in = 12
) {
  scaling_mode <- match.arg(scaling_mode)

  peak_rows <- as_tibble(peak_rows)
  peak_gr <- Signac::StringToGRanges(peak_rows$region, sep = c("-", "-"))
  chr <- as.character(seqnames(peak_gr))[1]
  pk_start <- min(start(peak_gr))
  pk_end <- max(end(peak_gr))
  pk_width <- pk_end - pk_start + 1
  if (is.null(fixed_window_bp)) {
    flank <- max(region_min_flank_bp, region_flank_factor * pk_width)
    from_bp <- max(1, pk_start - flank)
    to_bp <- pk_end + flank
  } else {
    half_window <- ceiling(fixed_window_bp / 2)
    peak_mid <- floor((pk_start + pk_end) / 2)
    from_bp <- max(1, peak_mid - half_window)
    to_bp <- peak_mid + half_window
  }

  gene_label <- unique(peak_rows$primary_gene[!is.na(peak_rows$primary_gene) & nzchar(peak_rows$primary_gene)])
  if (length(gene_label) == 0) gene_label <- unique(peak_rows$region)[1]
  gene_label <- gene_label[1]

  sig_list <- list()
  for (i in seq_len(nrow(track_specs))) {
    bw_vecs <- lapply(seq_along(track_specs$bw_paths[[i]]), function(j) {
      sid <- track_specs$sample_id[[i]][j]
      sf <- track_specs$size_factor[[i]][j]
      raw <- extract_region_signal(track_specs$bw_paths[[i]][j], chr, from_bp, to_bp)
      raw / sf
    })
    sig_mat <- do.call(rbind, bw_vecs)
    if (track_specs$aggregate_mode[i] == "mean") {
      sig <- colMeans(sig_mat, na.rm = TRUE)
    } else {
      sig <- sig_mat[1, ]
    }
    sig_list[[track_specs$track_name[i]]] <- sig
  }

  peak_mask <- rep(FALSE, length(sig_list[[1]]))
  for (j in seq_along(peak_gr)) {
    mask_start <- max(1, start(peak_gr)[j] - from_bp + 1)
    mask_end <- min(length(peak_mask), end(peak_gr)[j] - from_bp + 1)
    peak_mask[mask_start:mask_end] <- TRUE
  }

  display_sig_list <- lapply(sig_list, function(x) {
    smooth_signal(x, window_bp = smoothing_window_bp)
  })

  if (scaling_mode == "autoscaled") {
    display_sig_list <- lapply(display_sig_list, function(x) {
      qv <- max(x, na.rm = TRUE)
      if (!is.finite(qv) || qv <= 0) qv <- 1
      x / qv
    })
    scaled_vals <- unlist(display_sig_list, use.names = FALSE)
    ylim_top <- quantile(scaled_vals, probs = 0.995, na.rm = TRUE)
    if (!is.finite(ylim_top) || ylim_top <= 0) {
      ylim_top <- max(scaled_vals, na.rm = TRUE)
    }
    if (!is.finite(ylim_top) || ylim_top <= 0) ylim_top <- 1.05
    ylim_top <- max(1.02, min(1.12, ylim_top * 1.05))
  } else {
    non_epi_names <- setdiff(names(sig_list), "E6.5 Epiblast_C3")
    ref_vals <- unlist(display_sig_list[non_epi_names], use.names = FALSE)
    if (length(ref_vals) == 0) ref_vals <- unlist(display_sig_list, use.names = FALSE)
    ylim_top <- quantile(ref_vals, probs = 0.995, na.rm = TRUE)
    if (!is.finite(ylim_top) || ylim_top <= 0) {
      ylim_top <- max(ref_vals, na.rm = TRUE)
    }
    if (!is.finite(ylim_top) || ylim_top <= 0) ylim_top <- 1
  }

  track_display_names <- vapply(names(sig_list), simplify_track_label, character(1))
  pct_label_map <- NULL
  if (label_vs_epiblast && "E6.5 Epiblast_C3" %in% names(sig_list)) {
    epi_sig <- sig_list[["E6.5 Epiblast_C3"]]
    epi_mean <- mean(epi_sig[peak_mask], na.rm = TRUE)
    if (!is.finite(epi_mean) || epi_mean <= 0) epi_mean <- 1e-9

    track_display_names <- vapply(names(sig_list), function(nm) {
      if (nm == "E6.5 Epiblast_C3") {
        "C3"
      } else {
        simplify_track_label(nm)
      }
    }, character(1))

    pct_label_map <- vapply(names(sig_list), function(nm) {
      if (nm == "E6.5 Epiblast_C3") {
        NA_character_
      } else {
        this_mean <- mean(sig_list[[nm]][peak_mask], na.rm = TRUE)
        pct <- 100 * (this_mean - epi_mean) / epi_mean
        format_pct_delta(pct)
      }
    }, character(1))
    names(pct_label_map) <- vapply(names(sig_list), simplify_track_label, character(1))
  }

  gene_track <- GeneRegionTrack(
    txdb,
    chromosome = chr,
    start = from_bp,
    end = to_bp,
    collapseTranscripts = "meta",
    background.title = "white",
    name = ""
  )

  peak_track <- AnnotationTrack(
    start = start(peak_gr),
    end = end(peak_gr),
    chromosome = chr,
    genome = "mm10",
    name = "",
    fill = "#D95F02",
    col = "#D95F02"
  )

  displayPars(gene_track) <- list(
    background.title = "white",
    background.panel = "white",
    col.title = "black",
    col.axis = "black",
    cex.title = 0.7,
    fontsize = 10,
    fill = "#111111",
    col = "#111111",
    col.line = "#111111",
    just.group = "above",
    showId = FALSE,
    fontcolor.group = "#111111",
    lwd = 0.7,
    shape = "box"
  )
  displayPars(peak_track) <- list(background.title = "white", background.panel = "white", cex.title = 0.7)

  data_tracks <- list()
  for (k in seq_along(sig_list)) {
    nm <- names(sig_list)[k]
    this_color <- track_specs$track_color[match(nm, track_specs$track_name)]
    if (length(this_color) == 0 || is.na(this_color)) this_color <- "#4D4D4D"
    display_signal <- display_sig_list[[nm]]
    binned <- bin_signal(display_signal, chr, from_bp, bin_size = bin_size_local_bp)
    gr <- GRanges(seqnames = binned$seqnames, ranges = IRanges(start = binned$start, end = binned$end))
    tr <- DataTrack(
      range = gr,
      data = matrix(binned$signal, nrow = 1),
      genome = "mm10",
      chromosome = chr,
      type = "l",
      baseline = 0,
      name = track_display_names[k]
    )
    displayPars(tr) <- list(
      ylim = c(0, ylim_top),
      background.title = "white",
      background.panel = "white",
      col.axis = "black",
      col.title = "black",
      cex.axis = 0.75,
      cex.title = ifelse(compact, 0.62, 0.72),
      lwd = 1.2,
      col = this_color,
      lty = 1
    )
    data_tracks[[length(data_tracks) + 1]] <- tr
  }

  hl <- HighlightTrack(
    trackList = c(list(gene_track, peak_track), data_tracks),
    start = start(peak_gr),
    end = end(peak_gr),
    chromosome = chr,
    fill = peak_highlight_fill,
    col = NA,
    inBackground = TRUE
  )

  if (compact) {
    track_sizes <- c(0.18, 0.12, rep(0.42, length(data_tracks)))
    plot_height <- max(430, 110 + 62 * length(data_tracks))
  } else {
    track_sizes <- c(0.12, 0.07, rep(0.72, length(data_tracks)))
    plot_height <- max(500, 95 + 82 * length(data_tracks))
  }

  plot_gviz_core <- function() {
    plotTracks(
      hl,
      from = from_bp,
      to = to_bp,
      sizes = track_sizes,
      background.panel = "white",
      background.title = "white",
      col.axis = "black",
      cex.title = ifelse(compact, 0.58, 0.62),
      main = gene_label
    )
  }

  if (!is.null(out_png)) {
    png(out_png, width = png_width_px, height = plot_height, res = 220)
    plot_gviz_core()
    dev.off()
  }

  if (!is.null(out_svg)) {
    svglite::svglite(out_svg, width = svg_width_in, height = plot_height / 220)
    plot_gviz_core()
    dev.off()
    fix_svg_for_illustrator(out_svg)
    if (label_vs_epiblast && !is.null(pct_label_map)) {
      annotate_selected_panel_svg(
        svg_path = out_svg,
        peak_rows = peak_rows,
        from_bp = from_bp,
        to_bp = to_bp,
        pct_labels = pct_label_map,
        track_color_map = setNames(track_specs$track_color, vapply(track_specs$track_name, simplify_track_label, character(1)))
      )
    }
  }
}

make_track_specs <- function(study_meta, epiblast_meta, def, size_factor_tbl, mode = c("average", "replicates")) {
  mode <- match.arg(mode)

  if (mode == "average") {
    out <- list(
      tibble(
        track_name = "E6.5 Epiblast_C3",
        track_color = unname(study_track_colors["E6.5 Epiblast_C3"]),
        bw_paths = list(epiblast_meta$bw_path),
        sample_id = list(epiblast_meta$sample_id),
        size_factor = list(size_factor_tbl$size_factor[match(epiblast_meta$sample_id, size_factor_tbl$sample_id)]),
        aggregate_mode = "mean"
      )
    )

    for (grp in def$group_order) {
      samp <- study_meta %>% filter(sample_group == grp)
      out[[length(out) + 1]] <- tibble(
        track_name = grp,
        track_color = unname(study_track_colors[grp]),
        bw_paths = list(samp$bw_path),
        sample_id = list(samp$sample_id),
        size_factor = list(size_factor_tbl$size_factor[match(samp$sample_id, size_factor_tbl$sample_id)]),
        aggregate_mode = "mean"
      )
    }
    bind_rows(out)
  } else {
    epi_row <- tibble(
      track_name = "E6.5 Epiblast_C3",
      track_color = unname(study_track_colors["E6.5 Epiblast_C3"]),
      bw_paths = list(epiblast_meta$bw_path),
      sample_id = list(epiblast_meta$sample_id),
      size_factor = list(size_factor_tbl$size_factor[match(epiblast_meta$sample_id, size_factor_tbl$sample_id)]),
      aggregate_mode = "single"
    )

    rep_rows <- study_meta %>%
      mutate(track_name = sample_title, track_color = unname(study_track_colors[sample_group])) %>%
      transmute(
        track_name,
        track_color,
        bw_paths = purrr::map(bw_path, ~.x),
        sample_id = purrr::map(sample_id, ~.x),
        size_factor = purrr::map(sample_id, ~size_factor_tbl$size_factor[match(.x, size_factor_tbl$sample_id)]),
        aggregate_mode = "single"
      )

    bind_rows(epi_row, rep_rows)
  }
}

assemble_montage <- function(img_paths, title_text, subtitle_text, ncol = 3) {
  if (length(img_paths) == 0) {
    stop("No image paths were supplied to assemble_montage().")
  }
  plot_obj <- patchwork::wrap_plots(lapply(img_paths, make_image_plot), ncol = ncol)
  if (!nzchar(title_text) && !nzchar(subtitle_text)) {
    return(plot_obj)
  }
  plot_obj + plot_annotation(title = title_text, subtitle = subtitle_text)
}

parse_svg_dims <- function(svg_path) {
  doc <- xml2::read_xml(svg_path)
  root <- xml2::xml_root(doc)
  vb <- xml2::xml_attr(root, "viewBox")
  if (!is.na(vb) && nzchar(vb)) {
    nums <- as.numeric(strsplit(vb, "[ ,]+")[[1]])
    return(list(width = nums[3], height = nums[4]))
  }
  width_raw <- xml2::xml_attr(root, "width")
  height_raw <- xml2::xml_attr(root, "height")
  width_attr <- gsub("[^0-9.]", "", ifelse(is.na(width_raw), "", width_raw))
  height_attr <- gsub("[^0-9.]", "", ifelse(is.na(height_raw), "", height_raw))
  list(
    width = ifelse(nzchar(width_attr), as.numeric(width_attr), 1000),
    height = ifelse(nzchar(height_attr), as.numeric(height_attr), 800)
  )
}

prefix_svg_ids <- function(svg_text, prefix) {
  svg_text <- gsub('id="([^"]+)"', paste0('id="', prefix, '\\1"'), svg_text, perl = TRUE)
  svg_text <- gsub('url\\(#([^)]+)\\)', paste0('url(#', prefix, '\\1)'), svg_text, perl = TRUE)
  svg_text <- gsub('href="#([^"]+)"', paste0('href="#', prefix, '\\1"'), svg_text, perl = TRUE)
  svg_text <- gsub('xlink:href="#([^"]+)"', paste0('xlink:href="#', prefix, '\\1"'), svg_text, perl = TRUE)
  svg_text
}

fix_svg_for_illustrator <- function(svg_path) {
  doc <- xml2::read_xml(svg_path)
  root <- xml2::xml_root(doc)

  # Remove svglite CSS blocks and write base drawing defaults explicitly.
  xml2::xml_find_all(root, ".//*[local-name()='style']") %>% xml2::xml_remove()

  drawable_nodes <- xml2::xml_find_all(
    root,
    ".//*[local-name()='line' or local-name()='polyline' or local-name()='polygon' or local-name()='path' or local-name()='rect' or local-name()='circle']"
  )

  for (node in drawable_nodes) {
    if (is.na(xml2::xml_attr(node, "stroke-linecap"))) {
      xml2::xml_set_attr(node, "stroke-linecap", "round")
    }
    if (is.na(xml2::xml_attr(node, "stroke-linejoin"))) {
      xml2::xml_set_attr(node, "stroke-linejoin", "round")
    }
    if (is.na(xml2::xml_attr(node, "stroke-miterlimit"))) {
      xml2::xml_set_attr(node, "stroke-miterlimit", "10.00")
    }
  }

  text_nodes <- xml2::xml_find_all(root, ".//*[local-name()='text']")
  for (node in text_nodes) {
    if (is.na(xml2::xml_attr(node, "xml:space"))) {
      xml2::xml_set_attr(node, "xml:space", "preserve")
    }
  }

  poly_nodes <- xml2::xml_find_all(root, ".//*[local-name()='polyline']")
  for (node in poly_nodes) {
    pts <- xml2::xml_attr(node, "points")
    if (is.na(pts) || !nzchar(pts)) next
    pairs <- strsplit(trimws(pts), "\\s+")[[1]]

    style_attr <- xml2::xml_attr(node, "style")
    stroke_attr <- xml2::xml_attr(node, "stroke")
    stroke_col <- stroke_attr
    if ((is.na(stroke_col) || !nzchar(stroke_col)) && !is.na(style_attr) && grepl("stroke:", style_attr)) {
      stroke_col <- sub(".*stroke:\\s*([^;]+);?.*", "\\1", style_attr)
    }

    if (is.na(xml2::xml_attr(node, "fill")) || !nzchar(xml2::xml_attr(node, "fill"))) {
      xml2::xml_set_attr(node, "fill", "none")
    }
    if (!is.na(stroke_col) && nzchar(stroke_col) && !identical(stroke_col, "none")) {
      xml2::xml_set_attr(node, "stroke", stroke_col)
    }

    # Long polylines are the coverage traces; convert them to colored filled areas.
    if (length(pairs) < 100 || is.na(stroke_col) || !nzchar(stroke_col) || identical(stroke_col, "none")) next

    coords <- strsplit(pairs, ",")
    x_vals <- suppressWarnings(as.numeric(vapply(coords, `[`, character(1), 1)))
    y_vals <- suppressWarnings(as.numeric(vapply(coords, `[`, character(1), 2)))
    if (any(!is.finite(x_vals)) || any(!is.finite(y_vals))) next

    baseline_y <- max(y_vals, na.rm = TRUE)
    poly_pts <- paste(
      c(
        pairs,
        sprintf("%.2f,%.2f", x_vals[length(x_vals)], baseline_y),
        sprintf("%.2f,%.2f", x_vals[1], baseline_y)
      ),
      collapse = " "
    )

    area_node <- xml2::read_xml(
      sprintf(
        "<polygon points=\"%s\" fill=\"%s\" fill-opacity=\"1\" stroke=\"none\"/>",
        poly_pts, stroke_col
      )
    )
    xml2::xml_add_sibling(node, area_node, .where = "before")
  }

  glyph_paths <- xml2::xml_find_all(root, ".//*[local-name()='g' and contains(@class, 'glyphgroup')]//*[local-name()='path']")
  for (node in glyph_paths) {
    if (is.na(xml2::xml_attr(node, "stroke"))) {
      xml2::xml_set_attr(node, "stroke", "none")
    }
  }

  writeLines(as.character(root), con = svg_path)
  invisible(svg_path)
}

annotate_selected_panel_svg <- function(svg_path, peak_rows, from_bp, to_bp, pct_labels, track_color_map) {
  doc <- xml2::read_xml(svg_path)
  root <- xml2::xml_root(doc)

  # Simplify the main title to gene name only.
  title_node <- xml2::xml_find_first(root, ".//*[local-name()='text'][1]")
  if (!inherits(title_node, "xml_missing")) {
    gene_label <- unique(peak_rows$primary_gene[!is.na(peak_rows$primary_gene) & nzchar(peak_rows$primary_gene)])
    if (length(gene_label) == 0) gene_label <- unique(peak_rows$region)[1]
    xml2::xml_set_text(title_node, gene_label[1])
    xml2::xml_set_attr(title_node, "font-size", "16")
  }

  # Remove the grey genome-axis labels and subtitle-scale clutter if present.
  mb_nodes <- xml2::xml_find_all(root, ".//*[local-name()='text'][contains(text(),'mb')]")
  xml2::xml_remove(mb_nodes)

  grey_line_nodes <- xml2::xml_find_all(
    root,
    ".//*[local-name()='line' and contains(@style, '#A9A9A9')]"
  )
  xml2::xml_remove(grey_line_nodes)

  # Drop the generic subtitle text embedded in panel SVGs.
  all_text <- xml2::xml_find_all(root, ".//*[local-name()='text']")
  for (node in all_text) {
    txt <- xml2::xml_text(node)
    if (grepl("^Rows compare the same gene", txt)) {
      xml2::xml_remove(node)
    }
  }

  # Simplify left-side track labels and collect their centers.
  title_nodes <- xml2::xml_find_all(root, ".//*[local-name()='text'][contains(text(),'vs C3') or text()='E6.5 Epiblast_C3 (ref)' or text()='C3' or text()='ESC' or text()='EpiLC' or text()='EpiSC' or text()='sEpiLC' or text()='d2 EpiLC' or text()='EpiLC+CHIR' or text()='FTW' or text()='CHIR']")
  track_centers <- list()
  for (node in title_nodes) {
    txt <- xml2::xml_text(node)
    lbl <- gsub("\\s*\\([^)]*vs C3\\)", "", txt)
    lbl <- simplify_track_label(lbl)
    xml2::xml_set_text(node, lbl)
    xml2::xml_set_attr(node, "textLength", NULL)
    transform_txt <- xml2::xml_attr(node, "transform")
    if (!is.na(transform_txt) && grepl("translate\\(", transform_txt)) {
      vals <- strsplit(gsub(".*translate\\(([^)]*)\\).*", "\\1", transform_txt), ",")[[1]]
      if (length(vals) == 2) {
        track_centers[[lbl]] <- as.numeric(vals[2])
      }
    }
  }

  # Infer plot x-range from the widest plotting clip rectangle.
  plot_rects <- xml2::xml_find_all(
    root,
    ".//*[local-name()='clipPath']/*[local-name()='rect'][number(@width) > 200 and number(@height) > 20 and number(@height) < 80]"
  )
  if (length(plot_rects) == 0) {
    writeLines(as.character(root), con = svg_path)
    return(invisible(svg_path))
  }
  rect_widths <- suppressWarnings(as.numeric(xml2::xml_attr(plot_rects, "width")))
  plot_rect <- plot_rects[[which.max(rect_widths)]]
  plot_x <- as.numeric(xml2::xml_attr(plot_rect, "x"))
  plot_w <- as.numeric(xml2::xml_attr(plot_rect, "width"))

  peak_gr <- Signac::StringToGRanges(peak_rows$region, sep = c("-", "-"))
  peak_anchor_bp <- min(start(peak_gr)) - 0.10 * (max(end(peak_gr)) - min(start(peak_gr)))
  peak_anchor_bp <- max(from_bp, peak_anchor_bp)
  x_pct <- plot_x + ((peak_anchor_bp - from_bp) / (to_bp - from_bp)) * plot_w

  for (nm in names(pct_labels)) {
    if (!nm %in% names(track_centers)) next
    y_pct <- track_centers[[nm]] + 12
    fill_col <- ifelse(!is.na(track_color_map[nm]), track_color_map[nm], "black")
    xml2::xml_add_child(
      root,
      "text",
      pct_labels[[nm]],
      x = sprintf("%.2f", x_pct),
      y = sprintf("%.2f", y_pct),
      `font-size` = "9.5",
      `font-family` = "Arial",
      `font-weight` = "bold",
      fill = fill_col,
      `text-anchor` = "start",
      `xml:space` = "preserve"
    )
  }

  writeLines(as.character(root), con = svg_path)
  invisible(svg_path)
}

compose_svg_grid <- function(svg_paths, title_text, subtitle_text, ncol, out_svg) {
  if (length(svg_paths) == 0) {
    stop("No SVG paths were supplied to compose_svg_grid().")
  }

  svg_paths <- normalizePath(svg_paths, winslash = "/", mustWork = TRUE)
  dims <- parse_svg_dims(svg_paths[1])
  panel_width <- dims$width
  panel_height <- dims$height
  nrow_grid <- ceiling(length(svg_paths) / ncol)
  pad_x <- 18
  pad_y <- 18
  title_h <- if (nzchar(title_text)) 52 else 0
  subtitle_h <- if (nzchar(subtitle_text)) 34 else 0

  total_width <- ncol * panel_width + (ncol + 1) * pad_x
  total_height <- title_h + subtitle_h + nrow_grid * panel_height + (nrow_grid + 1) * pad_y

  root <- xml2::read_xml(
    sprintf(
      '<svg xmlns="http://www.w3.org/2000/svg" xmlns:xlink="http://www.w3.org/1999/xlink" width="%s" height="%s" viewBox="0 0 %s %s"></svg>',
      total_width, total_height, total_width, total_height
    )
  )

  xml2::xml_add_child(root, "rect", x = 0, y = 0, width = total_width, height = total_height, fill = "white")
  if (nzchar(title_text)) {
    xml2::xml_add_child(root, "text", title_text, x = pad_x, y = 30, `font-size` = 22, `font-family` = "Helvetica", `font-weight` = "600", fill = "black")
  }
  if (nzchar(subtitle_text)) {
    xml2::xml_add_child(root, "text", subtitle_text, x = pad_x, y = 56, `font-size` = 11, `font-family` = "Helvetica", fill = "black")
  }

  for (i in seq_along(svg_paths)) {
    row_idx <- (i - 1) %/% ncol
    col_idx <- (i - 1) %% ncol
    x_pos <- pad_x + col_idx * panel_width
    y_pos <- title_h + subtitle_h + pad_y + row_idx * (panel_height + pad_y)

    svg_text <- prefix_svg_ids(readr::read_file(svg_paths[i]), paste0("panel", i, "_"))
    panel_doc <- xml2::read_xml(svg_text)
    panel_root <- xml2::xml_root(panel_doc)
    view_box <- xml2::xml_attr(panel_root, "viewBox")
    if (is.na(view_box) || !nzchar(view_box)) {
      view_box <- paste(0, 0, panel_width, panel_height)
    }

    nested <- xml2::xml_add_child(
      root,
      "g",
      id = paste0("panel_group_", i),
      transform = paste0("translate(", x_pos, ",", y_pos, ")")
    )

    for (child in xml2::xml_children(panel_root)) {
      xml2::xml_add_child(nested, child, .copy = TRUE)
    }
  }

  writeLines(as.character(root), con = out_svg)
  fix_svg_for_illustrator(out_svg)
  invisible(out_svg)
}
```

## 4. Load Metadata and Marker Peaks

```{r load-data}
meta <- readr::read_delim(METADATA_FILE, delim = ";", show_col_types = FALSE) %>%
  mutate(
    sample_id = ifelse(is.na(basename) | basename == "", sub("[.]bw$", "", filename), basename),
    sample_title = clean_str(sample_title),
    bw_path = file.path(DATA_DIR, "ESC_other", filename),
    bw_exists = file.exists(bw_path)
  )

marker_tbl_all <- read.delim(MARKER_FILE, sep = "\t", stringsAsFactors = FALSE) %>%
  filter(avg_log2FC > marker_log2fc_cutoff, p_val_adj < marker_padj_cutoff) %>%
  mutate(primary_gene = harmonize_gene_symbols(extract_primary_gene(closest_TSS)))

marker_tbl_c3 <- marker_tbl_all %>%
  filter(cluster == "C3")

all_marker_gr <- Signac::StringToGRanges(marker_tbl_all$region, sep = c("-", "-"))
c3_marker_gr <- Signac::StringToGRanges(marker_tbl_c3$region, sep = c("-", "-"))

epiblast_meta <- tibble(
  sample_id = "Sendra_C3",
  sample_title = "E6.5 Epiblast_C3",
  bw_path = file.path(DATA_DIR, "Epiblast", "Sendra_C3.bw"),
  bw_exists = file.exists(bw_path)
) %>%
  filter(bw_exists)

study_definitions <- list(
  "PRJNA648362" = list(
    study_title = "PRJNA648362 | ESC to d2 EpiLC to EpiSC priming series",
    study_setup = paste(
      "Mouse ESC, day-2 EpiLC, and EpiSC chromatin profiles from a germline-competence study.",
      "The design compares discrete pluripotency states along the ESC -> EpiLC -> EpiSC progression."
    ),
    study_source = "Abe et al., Nat Commun 2021",
    study_source_url = "https://www.nature.com/articles/s41467-021-23639-8",
    comparison_a = "EpiSC",
    comparison_b = "ESC",
    group_order = c("ESC", "EpiLC", "EpiSC"),
    sample_map = c(
      "ATAC_ESC_rep1" = "ESC",
      "ATAC_ESC_rep2" = "ESC",
      "ATAC_EpiLC_rep1" = "EpiLC",
      "ATAC_EpiLC_rep2" = "EpiLC",
      "ATAC_EpiSC_rep1" = "EpiSC",
      "ATAC_EpiSC_rep2" = "EpiSC"
    )
  ),
  "PRJNA856446" = list(
    study_title = "PRJNA856446 | sEpiLC to d2 EpiLC to CHIR-supported FTW/EpiLSC series",
    study_setup = paste(
      "EpiLSC study profiling sequential formative-pluripotency states:",
      "sEpiLC, day-2 EpiLC, CHIR-treated EpiLC, and FTW/EpiLSC cells.",
      "CHIR is used to support the WNT-responsive formative state."
    ),
    study_source = "Kinoshita et al., Development 2023",
    study_source_url = "https://pubmed.ncbi.nlm.nih.gov/35946655/",
    comparison_a = "FTW",
    comparison_b = "sEpiLC",
    group_order = c("sEpiLC", "d2 EpiLC", "EpiLC+CHIR", "FTW"),
    sample_map = c(
      "sEpiLC_1" = "sEpiLC",
      "sEpiLC_2" = "sEpiLC",
      "d2 EpiLC_1" = "d2 EpiLC",
      "d2 EpiLC_2" = "d2 EpiLC",
      "EpiLC+CHIR_1" = "EpiLC+CHIR",
      "EpiLC+CHIR_2" = "EpiLC+CHIR",
      "FTW_1" = "FTW",
      "FTW_2" = "FTW"
    )
  )
)

study_meta_list <- lapply(names(study_definitions), function(study_id) {
  def <- study_definitions[[study_id]]
  meta %>%
    filter(study_accession == study_id, bw_exists, sample_title %in% names(def$sample_map)) %>%
    mutate(sample_group = unname(def$sample_map[sample_title])) %>%
    arrange(factor(sample_group, levels = def$group_order), sample_title)
})
names(study_meta_list) <- names(study_definitions)

cat("Strict markers all clusters:", nrow(marker_tbl_all), "\n")
cat("Strict C3 markers:", nrow(marker_tbl_c3), "\n")
```

## 5. Direct BigWig-Derived Normalization

```{r normalization, message=FALSE, warning=FALSE}
all_selected_samples <- bind_rows(study_meta_list) %>%
  distinct(sample_id, .keep_all = TRUE)

sample_signal_rows <- list()
for (i in seq_len(nrow(all_selected_samples))) {
  sid <- all_selected_samples$sample_id[i]
  bw <- all_selected_samples$bw_path[i]
  sample_signal_rows[[sid]] <- bw_mean_over_ranges(bw, all_marker_gr)
}

epiblast_all_marker <- bw_mean_over_ranges(epiblast_meta$bw_path[1], all_marker_gr)

norm_mat <- do.call(rbind, c(sample_signal_rows, list(Sendra_C3 = epiblast_all_marker)))
rownames(norm_mat) <- c(all_selected_samples$sample_id, "Sendra_C3")
colnames(norm_mat) <- marker_tbl_all$region

size_factor <- compute_size_factors(norm_mat)
size_factor_tbl <- tibble(sample_id = names(size_factor), size_factor = unname(size_factor))
readr::write_csv(size_factor_tbl, file.path(OUT_TABLES, "bigwig_direct_size_factors_selected_studies.csv"))
```

## 6. Rank C3 Peaks Per Study

```{r rank-peaks, message=FALSE, warning=FALSE}
c3_signal_rows <- list()
all_rank_rows <- list()
all_selected_rows <- list()
all_group_mean_rows <- list()
transition_count_rows <- list()
epiblast_count_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  peak_cols <- marker_tbl_c3$region

  sample_peak_means <- lapply(seq_len(nrow(sm)), function(i) {
    sid <- sm$sample_id[i]
    sf <- size_factor_tbl$size_factor[match(sid, size_factor_tbl$sample_id)]
    vals <- bw_mean_over_ranges(sm$bw_path[i], c3_marker_gr) / sf
    names(vals) <- peak_cols
    tibble(
      sample_id = sid,
      sample_title = sm$sample_title[i],
      sample_group = sm$sample_group[i]
    ) %>%
      bind_cols(as_tibble_row(as.list(vals)))
  })

  sample_peak_df <- bind_rows(sample_peak_means)
  peak_cols <- marker_tbl_c3$region

  mean_by_group <- sample_peak_df %>%
    group_by(sample_group) %>%
    summarise(across(all_of(peak_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop")

  epiblast_c3_vals <- bw_mean_over_ranges(epiblast_meta$bw_path[1], c3_marker_gr) /
    size_factor_tbl$size_factor[match(epiblast_meta$sample_id[1], size_factor_tbl$sample_id)]

  mean_long <- mean_by_group %>%
    pivot_longer(cols = all_of(peak_cols), names_to = "region", values_to = "mean_signal") %>%
    mutate(study_accession = study_id)
  all_group_mean_rows[[study_id]] <- mean_long

  vec_a <- as.numeric(mean_by_group[mean_by_group$sample_group == def$comparison_a, peak_cols, drop = TRUE])
  vec_b <- as.numeric(mean_by_group[mean_by_group$sample_group == def$comparison_b, peak_cols, drop = TRUE])

  rank_tbl <- marker_tbl_c3 %>%
    transmute(
      study_accession = study_id,
      primary_gene,
      region,
      avg_log2FC_marker = avg_log2FC,
      marker_padj = p_val_adj,
      mean_group_a = vec_a,
      mean_group_b = vec_b,
      log2FC_groupA_vs_groupB = log2((vec_a + 1e-3) / (vec_b + 1e-3)),
      delta_groupA_minus_groupB = vec_a - vec_b,
      higher_accessibility_condition = ifelse(log2FC_groupA_vs_groupB >= 0, def$comparison_a, def$comparison_b),
      lower_accessibility_condition = ifelse(log2FC_groupA_vs_groupB >= 0, def$comparison_b, def$comparison_a)
    ) %>%
    arrange(desc(log2FC_groupA_vs_groupB))

  readr::write_csv(rank_tbl, file.path(OUT_TABLES, paste0(study_id, "_ranked_C3_peaks.csv")))
  all_rank_rows[[study_id]] <- rank_tbl

  group_names <- def$group_order[def$group_order %in% mean_by_group$sample_group]
  if (length(group_names) >= 2) {
    contrast_pairs <- purrr::map(seq_len(length(group_names) - 1), ~c(group_names[.x], group_names[.x + 1]))
    contrast_pairs[[length(contrast_pairs) + 1]] <- c(def$comparison_b, def$comparison_a)

    for (cc in contrast_pairs) {
      x <- as.numeric(mean_by_group[match(cc[2], mean_by_group$sample_group), peak_cols, drop = TRUE])
      y <- as.numeric(mean_by_group[match(cc[1], mean_by_group$sample_group), peak_cols, drop = TRUE])
      log2fc_xy <- log2((x + 1e-3) / (y + 1e-3))
      category_xy <- classify_direction(log2fc_xy)

      transition_count_rows[[length(transition_count_rows) + 1]] <- tibble(
        study_accession = study_id,
        study_title = def$study_title,
        contrast = paste0(cc[1], "_to_", cc[2]),
        reference_from = cc[1],
        reference_to = cc[2],
        category = category_xy
      ) %>%
        count(study_accession, study_title, contrast, reference_from, reference_to, category, name = "n_peaks")
    }
  }

  for (grp in group_names) {
    grp_vals <- as.numeric(mean_by_group[match(grp, mean_by_group$sample_group), peak_cols, drop = TRUE])
    log2fc_epi <- log2((epiblast_c3_vals + 1e-3) / (grp_vals + 1e-3))
    category_epi <- dplyr::case_when(
      log2fc_epi >= stable_log2fc_threshold ~ "higher_in_Epiblast_C3",
      log2fc_epi <= -stable_log2fc_threshold ~ paste0("higher_in_", grp),
      TRUE ~ "similar"
    )
    epiblast_count_rows[[length(epiblast_count_rows) + 1]] <- tibble(
      study_accession = study_id,
      study_title = def$study_title,
      contrast = paste0("Epiblast_C3_vs_", grp),
      reference_from = grp,
      reference_to = "Epiblast_C3",
      category = category_epi
    ) %>%
      count(study_accession, study_title, contrast, reference_from, reference_to, category, name = "n_peaks")
  }

  selected_up <- rank_tbl %>%
    filter(log2FC_groupA_vs_groupB > 0) %>%
    arrange(desc(log2FC_groupA_vs_groupB), desc(mean_group_a)) %>%
    distinct_primary_gene() %>%
    slice_head(n = top_peaks_per_direction) %>%
    mutate(direction_class = "up")

  selected_down <- rank_tbl %>%
    filter(log2FC_groupA_vs_groupB < 0) %>%
    arrange(log2FC_groupA_vs_groupB, desc(mean_group_b)) %>%
    distinct_primary_gene() %>%
    slice_head(n = top_peaks_per_direction) %>%
    mutate(direction_class = "down")

  selected_tbl <- bind_rows(selected_up, selected_down) %>%
    mutate(
      study_title = def$study_title,
      comparison_label = paste0(def$comparison_a, " vs ", def$comparison_b)
    )

  readr::write_csv(selected_tbl, file.path(OUT_TABLES, paste0(study_id, "_selected_top_C3_peaks.csv")))
  all_selected_rows[[study_id]] <- selected_tbl
}

ranked_tbl_all <- bind_rows(all_rank_rows)
selected_tbl_all <- bind_rows(all_selected_rows)
group_means_tbl_all <- bind_rows(all_group_mean_rows)
transition_counts_tbl <- bind_rows(transition_count_rows)
epiblast_counts_tbl <- bind_rows(epiblast_count_rows)

readr::write_csv(ranked_tbl_all, file.path(OUT_TABLES, "all_selected_studies_ranked_C3_peaks.csv"))
readr::write_csv(selected_tbl_all, file.path(OUT_TABLES, "all_selected_studies_top_C3_peaks.csv"))
readr::write_csv(group_means_tbl_all, file.path(OUT_TABLES, "all_selected_studies_group_mean_C3_signal.csv"))
readr::write_csv(transition_counts_tbl, file.path(OUT_TABLES, "all_selected_studies_C3_transition_counts.csv"))
readr::write_csv(epiblast_counts_tbl, file.path(OUT_TABLES, "all_selected_studies_C3_vs_epiblast_counts.csv"))

selected_tbl_all
```

## 7. Study Context and C3 Peak Outcome Counts

The two study panels are meant to be self-explanatory, so the titles below summarize the biological setup that is being visualized.

- `PRJNA648362`: ESC, day-2 EpiLC, and EpiSC were profiled as discrete pluripotency states along the ESC to primed pluripotency progression in a germline-competence study. Source: [Abe et al., Nature Communications 2021](https://www.nature.com/articles/s41467-021-23639-8).
- `PRJNA856446`: sEpiLC, day-2 EpiLC, `EpiLC+CHIR`, and FTW/EpiLSC were profiled as a formative-pluripotency series, with CHIR supporting the WNT-responsive formative state. Source: [Kinoshita et al., Development 2023](https://pubmed.ncbi.nlm.nih.gov/35946655/).

```{r study-context-and-counts}
study_context_tbl <- tibble(
  study_accession = names(study_definitions),
  study_title = vapply(study_definitions, `[[`, character(1), "study_title"),
  experimental_setup = vapply(study_definitions, `[[`, character(1), "study_setup"),
  source = vapply(study_definitions, `[[`, character(1), "study_source"),
  source_url = vapply(study_definitions, `[[`, character(1), "study_source_url")
)

knitr::kable(study_context_tbl, align = "l")

cat("\nTransition counts across all strict C3 peaks (effect threshold: |log2FC| >=", round(stable_log2fc_threshold, 3), ", equivalent to >=1.2-fold):\n\n")
knitr::kable(
  transition_counts_tbl %>% arrange(study_accession, contrast, desc(category)),
  align = "l"
)

cat("\nC3 epiblast reference versus each study group:\n\n")
knitr::kable(
  epiblast_counts_tbl %>% arrange(study_accession, contrast, desc(category)),
  align = "l"
)
```

## 8. Render Real Browser Panels Per Locus

```{r render-locus-panels, message=FALSE, warning=FALSE}
render_manifest_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  sel_tbl <- all_selected_rows[[study_id]]

  avg_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "average")
  rep_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "replicates")

  study_dir <- file.path(OUT_LOCI, study_id)
  if (!dir.exists(study_dir)) dir.create(study_dir, recursive = TRUE)

  for (i in seq_len(nrow(sel_tbl))) {
    peak_row <- sel_tbl[i, ]
    gene_tag <- ifelse(is.na(peak_row$primary_gene) | peak_row$primary_gene == "", paste0("peak", i), peak_row$primary_gene)
    dir_tag <- peak_row$direction_class
    idx_tag <- sprintf("%02d", i)

    avg_abs_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_absolute.png"))
    avg_abs_svg <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_absolute.svg"))
    avg_auto_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_autoscaled.png"))
    avg_auto_svg <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_autoscaled.svg"))
    rep_abs_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_replicates_absolute.png"))
    rep_abs_svg <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_replicates_absolute.svg"))

    render_locus_gviz(
      peak_rows = peak_row,
      track_specs = avg_specs,
      scaling_mode = "absolute_normalized",
      out_png = avg_abs_png,
      out_svg = avg_abs_svg,
      title_prefix = paste0(study_id, " | average")
    )

    render_locus_gviz(
      peak_rows = peak_row,
      track_specs = avg_specs,
      scaling_mode = "autoscaled",
      out_png = avg_auto_png,
      out_svg = avg_auto_svg,
      title_prefix = paste0(study_id, " | average autoscaled")
    )

    render_locus_gviz(
      peak_rows = peak_row,
      track_specs = rep_specs,
      scaling_mode = "absolute_normalized",
      out_png = rep_abs_png,
      out_svg = rep_abs_svg,
      title_prefix = paste0(study_id, " | replicates")
    )

    render_manifest_rows[[length(render_manifest_rows) + 1]] <- tibble(
      study_accession = study_id,
      direction_class = peak_row$direction_class,
      primary_gene = peak_row$primary_gene,
      region = peak_row$region,
      average_absolute_png = avg_abs_png,
      average_absolute_svg = avg_abs_svg,
      average_autoscaled_png = avg_auto_png,
      average_autoscaled_svg = avg_auto_svg,
      replicates_absolute_png = rep_abs_png,
      replicates_absolute_svg = rep_abs_svg
    )
  }
}

render_manifest <- bind_rows(render_manifest_rows)
readr::write_csv(render_manifest, file.path(OUT_TABLES, "render_manifest_C3_browser_panels.csv"))
render_manifest
```

## 9. Selected Gene Locus Panels and Final Definitive Montage

```{r selected-gene-panels, message=FALSE, warning=FALSE}
selected_gene_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  avg_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "average")

  for (gene in selected_gene_priority) {
    gene_peaks <- ranked_tbl_all %>%
      filter(study_accession == study_id, primary_gene == gene) %>%
      arrange(desc(abs(log2FC_groupA_vs_groupB)), desc(abs(delta_groupA_minus_groupB)))

    if (nrow(gene_peaks) == 0) next

    gene_dir <- file.path(OUT_LOCI, study_id, "selected_gene_panels")
    if (!dir.exists(gene_dir)) dir.create(gene_dir, recursive = TRUE)

    out_png <- file.path(gene_dir, paste0(gene, "_selected_average_autoscaled.png"))
    out_svg <- file.path(gene_dir, paste0(gene, "_selected_average_autoscaled.svg"))

    tryCatch(
      {
        render_locus_gviz(
          peak_rows = gene_peaks,
          track_specs = avg_specs,
          scaling_mode = "autoscaled",
          out_png = out_png,
          out_svg = out_svg,
          title_prefix = def$study_title,
          compact = FALSE,
          label_vs_epiblast = TRUE,
          fixed_window_bp = selected_gene_fixed_window_bp,
          bin_size_local_bp = selected_gene_bin_size_bp,
          smoothing_window_bp = selected_gene_smoothing_window_bp,
          png_width_px = selected_gene_png_width_px,
          svg_width_in = selected_gene_svg_width_in
        )
      },
      error = function(e) {
        message("Fallback rendering for selected gene panel: ", study_id, " | ", gene, " | ", conditionMessage(e))
        render_locus_gviz(
          peak_rows = gene_peaks,
          track_specs = avg_specs,
          scaling_mode = "autoscaled",
          out_png = out_png,
          out_svg = out_svg,
          title_prefix = def$study_title,
          compact = FALSE,
          label_vs_epiblast = TRUE,
          fixed_window_bp = selected_gene_fixed_window_bp,
          bin_size_local_bp = selected_gene_bin_size_bp,
          smoothing_window_bp = selected_gene_smoothing_window_bp,
          png_width_px = ceiling(selected_gene_png_width_px * 1.2),
          svg_width_in = selected_gene_svg_width_in * 1.2
        )
      }
    )

    selected_gene_rows[[length(selected_gene_rows) + 1]] <- tibble(
      study_accession = study_id,
      study_title = def$study_title,
      gene = gene,
      n_c3_peaks_for_gene = nrow(gene_peaks),
      region_union = paste0(
        as.character(seqnames(Signac::StringToGRanges(gene_peaks$region[1], sep = c("-", "-"))))[1],
        ":",
        min(start(Signac::StringToGRanges(gene_peaks$region, sep = c("-", "-")))),
        "-",
        max(end(Signac::StringToGRanges(gene_peaks$region, sep = c("-", "-"))))
      ),
      panel_png = out_png,
      panel_svg = out_svg
    )
  }
}

selected_gene_manifest <- bind_rows(selected_gene_rows) %>%
  mutate(gene = factor(gene, levels = selected_gene_priority)) %>%
  arrange(gene, study_accession) %>%
  mutate(
    gene = as.character(gene),
    panel_png = normalizePath(panel_png, winslash = "/", mustWork = TRUE),
    panel_svg = normalizePath(panel_svg, winslash = "/", mustWork = TRUE)
  )

readr::write_csv(selected_gene_manifest, file.path(OUT_TABLES, "selected_gene_montage_manifest.csv"))
selected_gene_manifest
```

```{r selected-gene-definitive-montage, fig.width=13.5, fig.height=12, message=FALSE, warning=FALSE}
selected_gene_page_rows <- list()
gene_chunks <- split(selected_gene_priority, ceiling(seq_along(selected_gene_priority) / selected_gene_chunk_size))

for (page_idx in seq_along(gene_chunks)) {
  page_genes <- gene_chunks[[page_idx]]
  img_paths <- character()
  svg_paths <- character()
  for (gene_name in page_genes) {
    for (study_id in names(study_definitions)) {
      this_row <- selected_gene_manifest %>%
        filter(.data$study_accession == study_id, .data$gene == .env$gene_name) %>%
        slice_head(n = 1)
      if (nrow(this_row) == 1) {
        img_paths <- c(img_paths, this_row$panel_png)
        svg_paths <- c(svg_paths, this_row$panel_svg)
      }
    }
  }

  page_plot <- assemble_montage(
    img_paths = img_paths,
    title_text = "",
    subtitle_text = "",
    ncol = 2
  )

  page_png <- file.path(OUT_FIGS, paste0("selected_genes_definitive_montage_page_", sprintf("%02d", page_idx), ".png"))
  page_svg <- file.path(OUT_FIGS, paste0("selected_genes_definitive_montage_page_", sprintf("%02d", page_idx), ".svg"))

  ggsave(page_png, page_plot, width = 13.5, height = 12, dpi = 450)
  compose_svg_grid(
    svg_paths = svg_paths,
    title_text = "",
    subtitle_text = "",
    ncol = 2,
    out_svg = page_svg
  )

  selected_gene_page_rows[[length(selected_gene_page_rows) + 1]] <- tibble(
    page = page_idx,
    genes = paste(page_genes, collapse = ", "),
    montage_png = normalizePath(page_png, winslash = "/", mustWork = FALSE),
    montage_svg = normalizePath(page_svg, winslash = "/", mustWork = FALSE)
  )
}

selected_gene_pages <- bind_rows(selected_gene_page_rows)
readr::write_csv(selected_gene_pages, file.path(OUT_TABLES, "selected_genes_definitive_montage_pages.csv"))
selected_gene_pages
```

```{r selected-gene-montage-in-report, results='asis', echo=FALSE}
cat("\n\n### Final curated selected-gene montage\n\n")
for (i in seq_len(nrow(selected_gene_pages))) {
  cat("#### Montage page ", selected_gene_pages$page[i], " | genes: ", selected_gene_pages$genes[i], "\n\n", sep = "")
  print(knitr::include_graphics(selected_gene_pages$montage_png[i]))
  cat("\n\n")
}
```

## 10. Assemble Study Montages

```{r montage-assembly, fig.width=15, fig.height=10, message=FALSE, warning=FALSE}
study_montage_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  man <- render_manifest %>% filter(study_accession == study_id)
  if (nrow(man) == 0) next

  up_avg_abs <- man %>% filter(direction_class == "up") %>% pull(average_absolute_png)
  down_avg_abs <- man %>% filter(direction_class == "down") %>% pull(average_absolute_png)
  up_avg_auto <- man %>% filter(direction_class == "up") %>% pull(average_autoscaled_png)
  down_avg_auto <- man %>% filter(direction_class == "down") %>% pull(average_autoscaled_png)

  abs_plot <- (
    assemble_montage(
      up_avg_abs,
      title_text = paste0("More accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, library-size normalized"
    ) /
    assemble_montage(
      down_avg_abs,
      title_text = paste0("Less accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, library-size normalized"
    )
  ) + plot_annotation(
    title = def$study_title,
    subtitle = "Peak ranking and normalization derived directly from the BigWigs | y-limits set from non-epiblast tracks to avoid epiblast flattening"
  )

  auto_plot <- (
    assemble_montage(
      up_avg_auto,
      title_text = paste0("More accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, per-track autoscaled to the 99th percentile"
    ) /
    assemble_montage(
      down_avg_auto,
      title_text = paste0("Less accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, per-track autoscaled to the 99th percentile"
    )
  ) + plot_annotation(
    title = paste0(def$study_title, " | autoscaled"),
    subtitle = "Shape-comparison view only; amplitudes should not be interpreted quantitatively"
  )

  abs_png <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_absolute.png"))
  abs_svg <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_absolute.svg"))
  auto_png <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_autoscaled.png"))
  auto_svg <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_autoscaled.svg"))

  ggsave(abs_png, abs_plot, width = 15, height = 10, dpi = 450)
  ggsave(abs_svg, abs_plot, width = 15, height = 10, device = svglite::svglite)
  ggsave(auto_png, auto_plot, width = 15, height = 10, dpi = 450)
  ggsave(auto_svg, auto_plot, width = 15, height = 10, device = svglite::svglite)

  study_montage_rows[[length(study_montage_rows) + 1]] <- tibble(
    study_accession = study_id,
    study_title = def$study_title,
    average_absolute_png = abs_png,
    average_absolute_svg = abs_svg,
    average_autoscaled_png = auto_png,
    average_autoscaled_svg = auto_svg
  )
}

study_montage_manifest <- bind_rows(study_montage_rows)
readr::write_csv(study_montage_manifest, file.path(OUT_TABLES, "study_montage_manifest.csv"))
study_montage_manifest
```

## 11. All C3 Peaks, Autoscaled, Grouped in Montages of Six

```{r all-c3-autoscaled-compact, message=FALSE, warning=FALSE}
all_c3_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  avg_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "average")
  rank_tbl <- all_rank_rows[[study_id]] %>%
    arrange(desc(log2FC_groupA_vs_groupB), desc(abs(delta_groupA_minus_groupB))) %>%
    mutate(order_idx = row_number())

  compact_dir <- file.path(OUT_LOCI, study_id, "all_c3_autoscaled_compact")
  if (!dir.exists(compact_dir)) dir.create(compact_dir, recursive = TRUE)

  compact_paths <- character(nrow(rank_tbl))
  for (i in seq_len(nrow(rank_tbl))) {
    peak_row <- rank_tbl[i, ]
    gene_tag <- ifelse(is.na(peak_row$primary_gene) | peak_row$primary_gene == "", paste0("peak", i), peak_row$primary_gene)
    out_png <- file.path(compact_dir, paste0(sprintf("%03d", i), "_", gene_tag, "_autoscaled_compact.png"))
    compact_paths[i] <- out_png
    if (!file.exists(out_png)) {
      render_locus_gviz(
        peak_rows = peak_row,
        track_specs = avg_specs,
        scaling_mode = "autoscaled",
        out_png = out_png,
        title_prefix = paste0(study_id, " | compact autoscaled"),
        compact = TRUE
      )
    }
  }

  chunks <- split(compact_paths, ceiling(seq_along(compact_paths) / all_c3_chunk_size))
  for (page_idx in seq_along(chunks)) {
    montage_plot <- assemble_montage(
      chunks[[page_idx]],
      title_text = paste0(def$study_title, " | all strict C3 peaks | autoscaled"),
      subtitle_text = paste0(
        "Page ", page_idx, " of ", length(chunks),
        " | loci ordered by ", def$comparison_a, " vs ", def$comparison_b, " log2FC"
      ),
      ncol = 3
    )
    montage_png <- file.path(OUT_FIGS, paste0(study_id, "_all_C3_autoscaled_page_", sprintf("%02d", page_idx), ".png"))
    montage_svg <- file.path(OUT_FIGS, paste0(study_id, "_all_C3_autoscaled_page_", sprintf("%02d", page_idx), ".svg"))

    ggsave(montage_png, montage_plot, width = 15, height = 9, dpi = 350)
    ggsave(montage_svg, montage_plot, width = 15, height = 9, device = svglite::svglite)

    all_c3_rows[[length(all_c3_rows) + 1]] <- tibble(
      study_accession = study_id,
      study_title = def$study_title,
      page = page_idx,
      n_pages = length(chunks),
      montage_png = montage_png,
      montage_svg = montage_svg
    )
  }
}

all_c3_montage_manifest <- bind_rows(all_c3_rows)
readr::write_csv(all_c3_montage_manifest, file.path(OUT_TABLES, "all_C3_autoscaled_compact_montage_manifest.csv"))
all_c3_montage_manifest
```

## 12. All C3 Peak Montages in Report

```{r all-c3-in-report, results='asis', echo=FALSE}
for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sub_man <- all_c3_montage_manifest %>% filter(study_accession == study_id)
  cat("\n\n### ", def$study_title, " | all strict C3 peaks autoscaled\n\n", sep = "")
  for (i in seq_len(nrow(sub_man))) {
    cat("#### Montage page ", sub_man$page[i], " / ", sub_man$n_pages[i], "\n\n", sep = "")
    print(knitr::include_graphics(sub_man$montage_png[i]))
    cat("\n\n")
  }
}
```

## 13. Notes on Interpretation

```{r notes}
cat(
  paste(
    "- Main study figures use replicate averages because that is the clearest way to compare trajectory-level accessibility without overloading the panel.",
    "- Replicate-level browser plots are still saved for every selected locus in the locus_panels directory.",
    "- Real gene models come from Gviz + TxDb.Mmusculus.UCSC.mm10.knownGene.",
    "- Absolute normalized plots preserve between-group amplitude after size-factor normalization from the BigWigs.",
    "- Autoscaled plots divide each displayed track by its own 99th percentile and are meant only for shape comparison.",
    sep = "\n"
  )
)
```

## 14. Outputs

```{r outputs}
cat("Tables:\n")
cat("-", file.path(OUT_TABLES, "bigwig_direct_size_factors_selected_studies.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_ranked_C3_peaks.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_top_C3_peaks.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_group_mean_C3_signal.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_C3_transition_counts.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_C3_vs_epiblast_counts.csv"), "\n")
cat("-", file.path(OUT_TABLES, "render_manifest_C3_browser_panels.csv"), "\n")
cat("-", file.path(OUT_TABLES, "study_montage_manifest.csv"), "\n\n")
cat("-", file.path(OUT_TABLES, "all_C3_autoscaled_compact_montage_manifest.csv"), "\n\n")
cat("-", file.path(OUT_TABLES, "selected_gene_montage_manifest.csv"), "\n")
cat("-", file.path(OUT_TABLES, "selected_genes_definitive_montage_pages.csv"), "\n\n")

cat("Study montages:\n")
cat("-", file.path(OUT_FIGS, "PRJNA648362_C3_top_up_down_average_absolute.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA648362_C3_top_up_down_average_autoscaled.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA856446_C3_top_up_down_average_absolute.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA856446_C3_top_up_down_average_autoscaled.png"), "\n\n")
cat("Final selected-gene montage pages:\n")
cat("-", file.path(OUT_FIGS, "selected_genes_definitive_montage_page_01.png"), "\n")
cat("-", file.path(OUT_FIGS, "selected_genes_definitive_montage_page_02.png"), "\n\n")
cat("All C3 compact autoscaled montages:\n")
cat("-", OUT_FIGS, "\n\n")

cat("Per-locus panels:\n")
cat("-", OUT_LOCI, "\n")
```

## 15. Session Info

```{r session-info}
sessionInfo()
```
