---
title: "Tracking Behaviors Analysis"
author: "Refactored for reproducible execution"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    self_contained: false
    df_print: paged
---

# Setup

```{r}
source("R/behaviors_helpers.R")

install_missing_behavior_packages()
load_behavior_packages()

library(splines)
options(Seurat.object.assay.version = "v3")

rgl_available <- rgl_is_available()
if (rgl_available) {
  options(rgl.useNULL = TRUE, rgl.printRglwidget = TRUE)
  rgl::setupKnitr(autoprint = TRUE)
}

paths <- behavior_paths()
configure_knitr_for_behaviors(paths)

windows <- c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90, 100)
spline_df <- 5

analysis_state <- build_behavior_analysis_state(
  paths = paths,
  windows = windows,
  spline_df = spline_df
)
```

This notebook uses the locally available tracking inputs in `../data/`, including `KM1_tracking_extended_withDists.csv_alltracks.csv` for the expanded track table and `KM1_tracking.xlsx` for detailed track annotations.

# Input Data

```{r}
scrollable_kable(
  head(analysis_state$alltracks, 20),
  caption = paste0(
    "Input all-tracks table: ",
    nrow(analysis_state$alltracks),
    " rows x ",
    ncol(analysis_state$alltracks),
    " columns"
  ),
  digits = 4
)
```

```{r}
duplicate_summary <- data.frame(
  total_rows = nrow(analysis_state$alltracks),
  duplicated_node_time_rows = nrow(analysis_state$duplicate_resolution$duplicate_rows),
  corrected_rows = nrow(analysis_state$duplicate_resolution$corrected_tracks)
)

knitr::kable(duplicate_summary, caption = "Duplicate-resolution summary")
```

```{r}
scrollable_kable(
  head(analysis_state$duplicate_resolution$corrected_tracks, 25),
  caption = "Corrected track table after keeping the first record per node_id/timepoint",
  digits = 4
)
```

# Track Visualization

## 2D tracks

```{r fig.asp=1}
old_par <- par(no.readonly = TRUE)
par(mar = c(5, 4, 4, 8), xpd = TRUE, cex = 1)

plot(
  analysis_state$corrected_track_list$CM,
  main = "CM tracks",
  lwd = 1,
  cex = 0.25,
  pch.start = 1,
  pch.end = NULL,
  pch = 19,
  col = scales::alpha(as.factor(seq_along(analysis_state$corrected_track_list$CM)), 0.7)
)
legend(
  "topright",
  inset = c(-0.2, 0),
  legend = names(analysis_state$corrected_track_list$CM),
  pch = 19,
  col = as.factor(seq_along(analysis_state$corrected_track_list$CM))
)

par(mar = c(5, 4, 4, 8), xpd = TRUE, cex = 1)

plot(
  analysis_state$corrected_track_list$ET,
  main = "ET tracks",
  lwd = 1,
  cex = 0.25,
  pch.start = 1,
  pch.end = NULL,
  pch = 19,
  col = scales::alpha(as.factor(seq_along(analysis_state$corrected_track_list$ET)), 0.7)
)
legend(
  "topright",
  inset = c(-0.2, 0),
  legend = names(analysis_state$corrected_track_list$ET),
  pch = 19,
  col = as.factor(seq_along(analysis_state$corrected_track_list$ET))
)

par(old_par)
```

## 3D tracks

```{r webgl=TRUE, fig.asp=1}
if (rgl_available) {
  plot3d(analysis_state$corrected_track_list$CM, main = "CM tracks", tick.marks = FALSE)
  plot3d(analysis_state$corrected_track_list$ET, main = "ET tracks", tick.marks = FALSE)
} else {
  cat("3D track rendering was skipped because `rgl` could not load on this machine (missing X11/OpenGL runtime).")
}
```

# Track-Derived Measurements

```{r}
if (is.null(analysis_state$measurements_long)) {
  cat("This run uses the saved cleaned wide feature table from `../data/` to preserve the legacy results and keep render time practical on this machine.")
} else {
  scrollable_kable(
    head(analysis_state$measurements_long, 25),
    caption = paste0(
      "Track measurements by window: ",
      nrow(analysis_state$measurements_long),
      " rows x ",
      ncol(analysis_state$measurements_long),
      " columns"
    ),
    digits = 4
  )
}
```

```{r fig.asp=2.2}
metric_names <- unique(sub("_win\\d+$", "", names(analysis_state$measurements_wide)[-(1:4)]))

for (metric_name in metric_names) {
  metric_columns <- grep(paste0("^", metric_name, "_win"), names(analysis_state$measurements_wide), value = TRUE)
  plot_df <- analysis_state$measurements_wide |>
    dplyr::select(cell_type, node_id, timepoint, dplyr::all_of(metric_columns)) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(metric_columns),
      names_to = "metric_window",
      values_to = "value"
    ) |>
    dplyr::mutate(window = as.integer(sub(".*_win", "", metric_window)))

  mean_table <- plot_df |>
    dplyr::group_by(cell_type, timepoint, window) |>
    dplyr::summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

  plot_object <- ggplot(plot_df, aes(x = as.numeric(timepoint), colour = cell_type)) +
    geom_line(aes(y = value, group = node_id), alpha = 0.25) +
    geom_line(data = mean_table, aes(y = value), linewidth = 1.1) +
    facet_wrap(~ window, ncol = 2, scales = "free_y") +
    labs(title = metric_name, y = metric_name, x = "Timepoint") +
    theme_bw() +
    theme(legend.position = "top")

  print(plot_object)
}
```

## Wide table for longitudinal modeling

```{r}
scrollable_kable(
  head(analysis_state$measurements_wide, 20),
  caption = paste0(
    "Wide measurement table: ",
    nrow(analysis_state$measurements_wide),
    " rows x ",
    ncol(analysis_state$measurements_wide),
    " columns"
  ),
  digits = 4
)
```

# Longitudinal Mixed Models

```{r}
scrollable_kable(
  analysis_state$model_results,
  caption = "Longitudinal mixed-model summary with spline-adjusted time effects",
  digits = 4
)
```

```{r fig.width=4, fig.asp=5}
model_heatmap <- analysis_state$model_results
model_heatmap$Param2 <- paste(model_heatmap$Param, model_heatmap$Effect, sep = ".")
rownames(model_heatmap) <- model_heatmap$Param2

model_heatmap$signif_code <- ifelse(
  model_heatmap$pval.adj < 0.001, "***",
  ifelse(model_heatmap$pval.adj < 0.01, "**", ifelse(model_heatmap$pval.adj < 0.05, "*", " "))
)

heatmap_table <- model_heatmap |>
  dplyr::select(Param, Effect, Coef, signif_code) |>
  tidyr::pivot_wider(
    names_from = Effect,
    values_from = c(Coef, signif_code),
    names_sep = "."
  )

coef_columns <- grep("^Coef\\.", names(heatmap_table), value = TRUE)
coef_matrix <- as.matrix(heatmap_table[, coef_columns, drop = FALSE])
rownames(coef_matrix) <- heatmap_table$Param

colour_function <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

draw(
  Heatmap(
    coef_matrix,
    name = "Coefficient",
    col = colour_function,
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid::grid.text(
        heatmap_table[
          i,
          gsub(
            "^Coef\\.",
            "signif_code.",
            coef_columns[j]
          )
        ],
        x = x,
        y = y,
        gp = grid::gpar(fontsize = 10)
      )
    }
  )
)
```

# Wide-Table QC and Cleaning

```{r}
wide_description <- DataExplorer::introduce(analysis_state$measurements_wide)
knitr::kable(as.matrix(t(wide_description)), caption = "Summary of the raw wide table")
```

```{r fig.asp=0.4}
plot_intro(analysis_state$measurements_wide)
```

```{r fig.asp=4}
plot_missing(
  analysis_state$measurements_wide,
  geom_label_args = list(
    size = 2.5,
    label.padding = grid::unit(0.01, "lines"),
    hjust = -0.1,
    label.size = 0
  )
)
```

```{r}
cleaning_summary <- data.frame(
  raw_wide_rows = nrow(analysis_state$measurements_wide),
  raw_wide_cols = ncol(analysis_state$measurements_wide),
  clean_wide_rows = nrow(analysis_state$clean_wide),
  clean_wide_cols = ncol(analysis_state$clean_wide),
  clustering_rows = nrow(analysis_state$collapsed_duplicates$clustering_table),
  clustering_cols = ncol(analysis_state$collapsed_duplicates$clustering_table)
)

knitr::kable(cleaning_summary, caption = "Filtering and deduplication summary")
```

# Unsupervised Analysis

```{r}
colors_samples <- scales::hue_pal()(2)
colors_clusters <- c(
  RColorBrewer::brewer.pal(12, name = "Paired"),
  pals::glasbey(32)[-4],
  RColorBrewer::brewer.pal(9, name = "Set1"),
  RColorBrewer::brewer.pal(8, name = "Set2"),
  RColorBrewer::brewer.pal(12, name = "Set3"),
  RColorBrewer::brewer.pal(9, name = "Pastel1"),
  RColorBrewer::brewer.pal(8, name = "Pastel2"),
  RColorBrewer::brewer.pal(8, name = "Accent"),
  RColorBrewer::brewer.pal(8, name = "Dark2")
)
color_expression <- pals::jet(n = 50)
```

```{r}
clustering_table <- analysis_state$collapsed_duplicates$clustering_table
categorical_columns <- names(clustering_table)[1:4]
feature_columns <- names(clustering_table)[5:ncol(clustering_table)]

seurat_meta <- clustering_table[, categorical_columns, drop = FALSE]
seurat_meta$node_id <- as.numeric(as.character(seurat_meta$node_id))
seurat_meta$timepoint <- as.numeric(as.character(seurat_meta$timepoint))

corrected_tracks <- analysis_state$duplicate_resolution$corrected_tracks
seurat_meta <- merge(seurat_meta, corrected_tracks, all.x = TRUE, all.y = FALSE)

tracking_matrix <- t(as.matrix(clustering_table[, feature_columns, drop = FALSE]))
colnames(tracking_matrix) <- rownames(clustering_table)

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_pca.rds"),
  function() {
    object <- suppressWarnings(
      CreateSeuratObject(
        counts = tracking_matrix,
        assay = "IM",
        meta.data = seurat_meta,
        project = "Tracking"
      )
    )
    DefaultAssay(object) <- "IM"
    VariableFeatures(object) <- rownames(object[["IM"]])
    object |>
      ScaleData(verbose = FALSE) |>
      RunPCA(verbose = FALSE)
  }
)
```

## PCA and clustering

```{r}
std_pc <- Stdev(SO, reduction = "pca")
elbow_point <- estimate_elbow_point(std_pc)
pcas <- seq_len(elbow_point)

ggplot(data.frame(PC = seq_along(std_pc), STD = std_pc), aes(x = PC, y = STD)) +
  geom_point() +
  geom_vline(xintercept = elbow_point, colour = "red") +
  ggtitle(paste("Elbow point:", elbow_point)) +
  theme_bw()
```

```{r}
use_assay <- "IM"
resolutions <- c(0.1, 0.2, 0.5, 0.75)
clustering_names <- paste0(use_assay, "_snn_res.", resolutions)

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_clusters.rds"),
  function() {
    clustered <- FindNeighbors(SO, dims = pcas, reduction = "pca")
    clustered <- FindClusters(clustered, resolution = resolutions, verbose = FALSE)

    for (cluster_name in clustering_names) {
      clustered[[cluster_name]] <- as.factor(
        paste0("C", stringr::str_pad(clustered[[cluster_name]][, 1], 2, pad = "0"))
      )
    }

    clustered
  }
)

cluster_counts <- SO@meta.data[, clustering_names, drop = FALSE] |>
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "resolution", values_to = "cluster") |>
  dplyr::count(resolution, cluster, name = "n_cells")

knitr::kable(cluster_counts, caption = "Cluster sizes across tested resolutions")
```

## tSNE and UMAP

```{r}
selected_resolution <- 0.5
resolution_name <- paste0(use_assay, "_snn_res.", selected_resolution)
SO <- SetIdent(SO, value = resolution_name)
DefaultAssay(SO) <- use_assay

selected_perplexity <- 200
selected_neighbors <- 50
selected_min_distance <- 0.3

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_embeddings_final.rds"),
  function() {
    object <- SO
    object <- RunTSNE(
      object,
      assay = use_assay,
      dims = pcas,
      seed.use = 123456,
      perplexity = selected_perplexity,
      check_duplicates = FALSE,
      reduction.name = "tSNE.Final",
      reduction.key = "tSNE_",
      verbose = FALSE
    )
    object <- RunUMAP(
      object,
      dims = pcas,
      seed.use = 123456,
      n.neighbors = selected_neighbors,
      min.dist = selected_min_distance,
      reduction.name = "UMAP.Final",
      reduction.key = "UMAP_",
      verbose = FALSE
    )
    object
  }
)
```

```{r fig.width=12, fig.asp=0.35}
SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_final.rds"),
  function() {
    final_object <- SetIdent(SO, value = resolution_name)
    final_object$res.Final <- final_object[[resolution_name]][, 1]
    SetIdent(final_object, value = "res.Final")
  }
)

SO <- add_annotations_to_seurat(SO, analysis_state$node_mapping, ntimepoints = 5)

DimPlot(
  SO,
  reduction = "tSNE.Final",
  group.by = "cell_type",
  cols = colors_samples,
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "tSNE.Final",
    group.by = "res.Final",
    cols = colors_clusters,
    shuffle = TRUE
  )
```

```{r fig.width=12, fig.asp=0.35}
DimPlot(
  SO,
  reduction = "UMAP.Final",
  group.by = "cell_type",
  cols = colors_samples,
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "UMAP.Final",
    group.by = "res.Final",
    cols = colors_clusters,
    shuffle = TRUE
  )
```

```{r fig.width=12, fig.asp=0.4}
FeaturePlot(
  SO,
  reduction = "UMAP.Final",
  features = c("node_id2", "ntimepoint"),
  combine = TRUE
) &
  scale_colour_gradientn(colors = color_expression, na.value = "grey90")
```

```{r fig.width=12, fig.asp=0.4}
selected_features <- c(
  "trackLength_win1",
  "maxDisplacement_win20",
  "straightness_win20",
  "asphericity_win20",
  "DistBorderMean_win20",
  "overallAngle_win20"
)
selected_features <- gsub("_", "-", selected_features, fixed = TRUE)

FeaturePlot(
  SO,
  reduction = "UMAP.Final",
  features = selected_features[selected_features %in% VariableFeatures(SO)],
  combine = TRUE
) &
  scale_colour_gradientn(colors = color_expression, na.value = "grey90")
```

```{r fig.width=12, fig.asp=0.6}
DimPlot(
  SO,
  reduction = "UMAP.Final",
  group.by = "cell_type_detailed",
  cols = RColorBrewer::brewer.pal(8, "Set1"),
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "UMAP.Final",
    group.by = "Embryo",
    cols = RColorBrewer::brewer.pal(8, "Set2"),
    shuffle = TRUE
  )
```

# Reference Comparison

```{r}
reference_checks <- dplyr::bind_rows(
  cbind(
    artifact = "clean_wide_nodups",
    compare_with_reference(
      analysis_state$collapsed_duplicates$clustering_table,
      paths$legacy_clean_wide
    )
  ),
  cbind(
    artifact = "longitudinal_stats",
    compare_with_reference(
      analysis_state$model_results,
      paths$legacy_stats
    )
  )
)

knitr::kable(reference_checks, caption = "Structural comparison against legacy reference files")
```

# Output Files

Generated tables and caches are written to `r paths$output_root`.

```{r}
generated_files <- data.frame(
  file = sort(list.files(paths$output_root, recursive = TRUE))
)

knitr::kable(generated_files, caption = "Files generated by the refactored notebook")
```

# Session Info

```{r}
sessionInfo()
```
