---
title: "Tracking Behaviors Analysis: Explainer Version"
author: "Refactored for lab review"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: united
    self_contained: false
    df_print: paged
---

# Purpose

This notebook is the review-friendly companion to `Behaviors_clean.Rmd`. It runs the same analysis but slows down the explanation. The goal is to make it easy to answer questions such as:

- What is the input table, and why is it already in an "all tracks" format?
- Why are duplicated `(node_id, timepoint)` rows removed before measurements?
- What does each temporal window represent biologically and computationally?
- Why is the mixed model fit on the wide table with missing values still present?
- Why are the clustering steps based on a later cleaned and deduplicated table?

> Review note  
> The original `Behaviors_v16.Rmd` expected inputs on a Windows network drive. In this workspace, those relevant inputs are available locally in `../data/`, including the expanded all-tracks CSV and the raw workbook `KM1_tracking.xlsx`. The refactor keeps the same analysis flow while replacing machine-specific paths with local relative paths.

# Environment and Reproducibility

## Package bootstrap

```{r}
source("R/behaviors_helpers.R")

install_missing_behavior_packages()
load_behavior_packages()

library(splines)
options(Seurat.object.assay.version = "v3")

rgl_available <- rgl_is_available()
if (rgl_available) {
  options(rgl.useNULL = TRUE, rgl.printRglwidget = TRUE)
  rgl::setupKnitr(autoprint = TRUE)
}
```

Why this chunk exists:

- `install_missing_behavior_packages()` makes the notebook self-starting on a fresh R installation.
- `load_behavior_packages()` keeps the rest of the notebook readable by loading the required packages in one place.
- `splines` is loaded explicitly because the longitudinal model uses `ns(timepoint, 5)`.
- `rgl.useNULL = TRUE` allows 3D plotting while knitting HTML in a non-interactive session.

## Path configuration

```{r}
paths <- behavior_paths()
configure_knitr_for_behaviors(paths)

windows <- c(1, 5, 10, 15, 20, 25, 30, 35, 40, 45, 50, 60, 70, 80, 90, 100)
spline_df <- 5

knitr::kable(
  data.frame(
    item = c("analysis_dir", "data_dir", "output_root", "cache_dir", "alltracks_csv", "track_details_xlsx"),
    path = c(
      paths$analysis_dir,
      paths$data_dir,
      paths$output_root,
      paths$cache_dir,
      paths$alltracks_csv,
      paths$track_details_xlsx
    )
  ),
  caption = "Key project paths used by the refactored notebook"
)
```

Why this matters:

- The original notebook wrote results to `S:/...`, which made it non-portable.
- The refactor assumes execution starts inside `figures_scripts`, then derives all paths relative to that directory.
- Generated CSVs, RDS caches, and plots are kept under `figures_scripts/output/behaviors_v16`.

# Build the Shared Analysis State

```{r}
analysis_state <- build_behavior_analysis_state(
  paths = paths,
  windows = windows,
  spline_df = spline_df
)
```

This helper does several things in a fixed order:

1. Reads the all-tracks CSV.
2. Reads `KM1_tracking.xlsx` so the final object can recover detailed annotations such as `cell_type_detailed`.
3. Builds `celltrackR` track objects from the positional coordinates.
4. Identifies duplicated `(node_id, timepoint)` rows and keeps the first record.
5. Rebuilds the corrected track objects.
6. Computes track-derived measurements for each temporal window.
7. Converts the long measurement table into a wide cell-timepoint feature table.
8. Fits the longitudinal mixed models.
9. Creates the cleaned and deduplicated clustering input.
10. Builds a node-ID correspondence table so the final Seurat object can still be traced back to the original track identities.

> Review note  
> The original notebook referenced a missing helper script named `LinearMixModel_UnivDJC_nowarn_v2.R` and also used undefined objects `dt.ord2` and `long.vars`. In the refactor, that step is replaced by a local function `fit_longitudinal_models()`, which reconstructs the intended behavior directly from the wide table.

# Data Loading and Early QC

## What the input table contains

```{r}
scrollable_kable(
  head(analysis_state$alltracks, 20),
  caption = paste0(
    "Input all-tracks table: ",
    nrow(analysis_state$alltracks),
    " rows x ",
    ncol(analysis_state$alltracks),
    " columns"
  ),
  digits = 4
)
```

Conceptually:

- Each row is one cell observation at one timepoint.
- `node_id` identifies the terminal track lineage used in the expanded all-tracks table.
- `track_name` preserves the original label from the manual or semi-manual tracking workflow.
- `X`, `Y`, and `Z` are the spatial coordinates.
- `DistBorder` is carried into the later feature engineering step through rolling means.

## Duplicate `(node_id, timepoint)` rows

```{r}
duplicate_overview <- data.frame(
  total_rows = nrow(analysis_state$alltracks),
  duplicated_node_time_rows = nrow(analysis_state$duplicate_resolution$duplicate_rows),
  corrected_rows = nrow(analysis_state$duplicate_resolution$corrected_tracks)
)

knitr::kable(duplicate_overview, caption = "Duplicate-resolution overview")
```

```{r}
scrollable_kable(
  head(analysis_state$duplicate_resolution$duplicate_rows, 20),
  caption = "Example rows that duplicated the same node_id/timepoint combination",
  digits = 4
)
```

Why duplicates matter:

- `celltrackR::as.tracks()` expects one position per track and timepoint.
- If the same `node_id` has two positions at the same timepoint, trajectory-derived quantities become ambiguous.
- The original notebook resolved this by keeping the first instance. The refactor keeps the same rule and documents it explicitly.

# Track Representation

## 2D visualization

```{r fig.asp=1}
old_par <- par(no.readonly = TRUE)
par(mar = c(5, 4, 4, 8), xpd = TRUE, cex = 1)

plot(
  analysis_state$corrected_track_list$CM,
  main = "CM tracks",
  lwd = 1,
  cex = 0.25,
  pch.start = 1,
  pch.end = NULL,
  pch = 19,
  col = scales::alpha(as.factor(seq_along(analysis_state$corrected_track_list$CM)), 0.7)
)

plot(
  analysis_state$corrected_track_list$ET,
  main = "ET tracks",
  lwd = 1,
  cex = 0.25,
  pch.start = 1,
  pch.end = NULL,
  pch = 19,
  col = scales::alpha(as.factor(seq_along(analysis_state$corrected_track_list$ET)), 0.7)
)

par(old_par)
```

Interpretation:

- These plots give a qualitative sense of cell displacement patterns before reducing the data into summary features.
- They are useful for checking gross artifacts such as obvious coordinate inversions, truncated tracks, or one cell type having systematically shorter trajectories due to preprocessing rather than biology.

## 3D visualization

```{r webgl=TRUE, fig.asp=1}
if (rgl_available) {
  plot3d(analysis_state$corrected_track_list$CM, main = "CM tracks", tick.marks = FALSE)
  plot3d(analysis_state$corrected_track_list$ET, main = "ET tracks", tick.marks = FALSE)
} else {
  cat("3D track rendering was skipped because `rgl` could not load on this machine (missing X11/OpenGL runtime).")
}
```

Why keep 3D plots in the review notebook:

- The measurements are computed from full trajectories, not from 2D projections.
- Showing the 3D object reminds the audience that quantities such as displacement, straightness, and asphericity are spatial features, not image-plane features.

# Feature Engineering

## What the temporal windows mean

The notebook computes behavior metrics over windows of length:

```{r}
knitr::kable(data.frame(window = windows), caption = "Temporal windows used for subtrack measurements")
```

How to explain this in a review:

- A window of `1` means "use very local motion behavior".
- A window of `20` means "summarize the subtrack spanning 20 steps".
- Larger windows smooth short-term fluctuations but reduce the number of eligible starting timepoints.
- This is why missingness increases with larger windows and why the later wide table needs explicit filtering.

## Long-format measurement table

```{r}
if (is.null(analysis_state$measurements_long)) {
  cat("The refactored notebook is using the saved cleaned wide feature table from `../data/` instead of recomputing every windowed measurement during render. This keeps the lab-review notebook reproducible and fast on the current machine while preserving the downstream legacy results.")
} else {
  scrollable_kable(
    head(analysis_state$measurements_long, 25),
    caption = "Long-format table: one row per cell, starting timepoint, and window",
    digits = 4
  )
}
```

The key columns are:

- `window`: which temporal scale was used.
- `node_id`: which trajectory the subtrack belongs to.
- `timepoint`: the starting timepoint of that subtrack.
- `t`: the subtrack index returned by `celltrackR`.
- The remaining columns are trajectory summaries such as `trackLength`, `straightness`, `overallAngle`, and `DistBorderMean`.

## Mean curves by cell type

```{r fig.asp=2.2}
metric_names <- unique(sub("_win\\d+$", "", names(analysis_state$measurements_wide)[-(1:4)]))

for (metric_name in metric_names) {
  metric_columns <- grep(paste0("^", metric_name, "_win"), names(analysis_state$measurements_wide), value = TRUE)
  plot_df <- analysis_state$measurements_wide |>
    dplyr::select(cell_type, node_id, timepoint, dplyr::all_of(metric_columns)) |>
    tidyr::pivot_longer(
      cols = dplyr::all_of(metric_columns),
      names_to = "metric_window",
      values_to = "value"
    ) |>
    dplyr::mutate(window = as.integer(sub(".*_win", "", metric_window)))

  mean_table <- plot_df |>
    dplyr::group_by(cell_type, timepoint, window) |>
    dplyr::summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

  ggplot(plot_df, aes(x = as.numeric(timepoint), colour = cell_type)) +
    geom_line(aes(y = value, group = node_id), alpha = 0.25) +
    geom_line(data = mean_table, aes(y = value), linewidth = 1.1) +
    facet_wrap(~ window, ncol = 2, scales = "free_y") +
    labs(title = metric_name, y = metric_name, x = "Timepoint") +
    theme_bw() +
    theme(legend.position = "top") |>
    print()
}
```

How to read these plots:

- Thin lines show individual-cell trajectories.
- Thick lines show the mean trend per cell type.
- Comparing panels across windows reveals whether a signal is robust across short and long temporal scales.

# Why There Are Two Wide Tables

## Wide table used for modeling

```{r}
scrollable_kable(
  head(analysis_state$measurements_wide, 20),
  caption = "Wide table used for longitudinal modeling",
  digits = 4
)
```

Important review point:

- The mixed models are fit on this table before complete-case filtering.
- That preserves many more observations for short windows.
- This choice explains why the `Nobs` values in the model output are much larger than the row count in the later clustering table.

## Wide table used for clustering

```{r}
cleaning_summary <- data.frame(
  raw_wide_rows = nrow(analysis_state$measurements_wide),
  raw_wide_cols = ncol(analysis_state$measurements_wide),
  clean_wide_rows = nrow(analysis_state$clean_wide),
  clean_wide_cols = ncol(analysis_state$clean_wide),
  clustering_rows = nrow(analysis_state$collapsed_duplicates$clustering_table),
  clustering_cols = ncol(analysis_state$collapsed_duplicates$clustering_table)
)

knitr::kable(cleaning_summary, caption = "Why the clustering table is smaller")
```

Why the second filtering stage exists:

- PCA, neighbor graph construction, tSNE, and UMAP do not behave well with many missing values.
- The notebook first removes all-empty columns.
- It then removes columns with more than 16% missingness, which effectively discards the longest windows.
- Finally it keeps complete cases and collapses feature-identical duplicate rows to avoid downstream issues in Seurat.

## Dataset structure diagnostics

```{r}
wide_description <- DataExplorer::introduce(analysis_state$measurements_wide)
knitr::kable(as.matrix(t(wide_description)), caption = "High-level description of the raw wide table")
```

```{r fig.asp=0.4}
plot_intro(analysis_state$measurements_wide)
```

```{r fig.asp=4}
plot_missing(
  analysis_state$measurements_wide,
  geom_label_args = list(
    size = 2.5,
    label.padding = grid::unit(0.01, "lines"),
    hjust = -0.1,
    label.size = 0
  )
)
```

# Longitudinal Statistical Model

## Why a mixed-effects model is the right tool here

The central statistical issue is that these rows are **repeated measurements from the same tracked cells**.

- One `node_id` contributes many rows over time.
- Rows from the same cell are more similar to each other than rows from different cells.
- The time trend is visibly curved rather than linear.

So we need a model that can do two things simultaneously:

1. Estimate the **average biological pattern** across cell types and time.
2. Respect the fact that **each cell has its own trajectory**, so repeated observations are not independent.

If we ignored that structure and used an ordinary linear model treating every row as independent:

- uncertainty would usually be underestimated,
- p-values could look more confident than they should,
- and cell-to-cell heterogeneity would leak into the group comparison.

That is exactly what the mixed-effects model protects us from.

## The model being fit

For each feature column in the wide table, the refactor fits:

```text
response_value ~ cell_type * ns(timepoint, 5) + (1 + timepoint || node_id)
```

How to explain each term:

- `response_value`: the current behavioral feature, standardized to mean 0 and variance 1.
- `cell_type`: tests the baseline difference between ET and the CM reference group.
- `ns(timepoint, 5)`: allows a smooth but non-linear time effect with 5 spline basis functions.
- `cell_type * ns(timepoint, 5)`: asks whether the time trajectory differs between cell types.
- `(1 + timepoint || node_id)`: gives each tracked cell its own intercept and time slope, with uncorrelated random effects.

## Professor-style intuition

At a high level, the model splits the story into two layers:

- **fixed effects**: the average CM baseline, the average ET-vs-CM difference, and the average time trend,
- **random effects**: cell-specific departures from that average pattern.

### Schematic 1: repeated measures are nested within cells

```{r fig.width=11, fig.height=4.8}
set.seed(42)

toy_cells <- paste0("Cell_", seq_len(10))
toy_time <- rep(seq(0, 9), times = length(toy_cells))
toy_group <- rep(rep(c("CM", "ET"), each = 5), each = 10)
toy_cell <- rep(toy_cells, each = 10)

random_intercept <- rep(rnorm(length(toy_cells), sd = 0.7), each = 10)
random_slope <- rep(rnorm(length(toy_cells), sd = 0.12), each = 10)
group_shift <- ifelse(toy_group == "ET", 0.7, 0)

toy_df <- data.frame(
  cell = toy_cell,
  cell_type = toy_group,
  timepoint = toy_time
) |>
  dplyr::mutate(
    response = 2 + 0.18 * timepoint - 0.02 * timepoint^2 + group_shift +
      random_intercept + random_slope * timepoint + rnorm(dplyr::n(), sd = 0.18)
  )

toy_mean <- toy_df |>
  dplyr::group_by(cell_type, timepoint) |>
  dplyr::summarise(response = mean(response), .groups = "drop")

ggplot(toy_df, aes(timepoint, response, group = cell, color = cell_type)) +
  geom_line(alpha = 0.35, linewidth = 0.5) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_line(data = toy_mean, aes(group = cell_type), linewidth = 1.4) +
  facet_wrap(~ cell_type) +
  theme_bw() +
  labs(
    title = "Toy example: individual cells fluctuate around a group-average trajectory",
    subtitle = "Thin lines are individual cells; thick lines are the average pattern the fixed effects try to capture",
    x = "Timepoint",
    y = "Example feature value"
  )
```

How to explain this figure:

- The thin lines are the actual repeated measurements from individual cells.
- The thick line is the group-average trajectory.
- The mixed model learns the thick-line behavior while acknowledging that the thin lines from the same cell are correlated.

### Schematic 2: what the model is combining

```{r fig.width=11, fig.height=5}
schematic_df <- tibble::tribble(
  ~component, ~x, ~y, ~label,
  "Observed rows", 1.1, 3.0, "Observed repeated rows\n(cell-timepoint measurements)",
  "Fixed effects", 4.0, 4.0, "Fixed effects\naverage CM baseline\naverage ET-CM contrast\naverage time trend",
  "Random effects", 4.0, 2.0, "Random effects by node_id\ncell-specific intercept\ncell-specific slope",
  "Inference", 7.2, 3.0, "Model output\ncoefficients\nconfidence intervals\np-values"
)

arrow_df <- tibble::tribble(
  ~x, ~y, ~xend, ~yend,
  2.0, 3.0, 3.2, 3.8,
  2.0, 3.0, 3.2, 2.2,
  4.9, 4.0, 6.5, 3.2,
  4.9, 2.0, 6.5, 2.8
)

ggplot() +
  geom_curve(
    data = arrow_df,
    aes(x = x, y = y, xend = xend, yend = yend),
    curvature = 0.08,
    linewidth = 0.6,
    color = "grey35",
    arrow = grid::arrow(length = grid::unit(0.18, "cm"))
  ) +
  geom_label(
    data = schematic_df,
    aes(x = x, y = y, label = label, fill = component),
    label.size = 0.25,
    size = 4,
    label.padding = grid::unit(0.3, "lines"),
    color = "black"
  ) +
  scale_fill_manual(
    values = c(
      "Observed rows" = "#d9ead3",
      "Fixed effects" = "#f9d5a7",
      "Random effects" = "#cfe2f3",
      "Inference" = "#ead1dc"
    )
  ) +
  coord_cartesian(xlim = c(0, 8.2), ylim = c(1.1, 4.9), clip = "off") +
  theme_void() +
  theme(legend.position = "none") +
  labs(title = "Conceptual anatomy of the mixed-effects model")
```

This is the punchline:

- the **fixed effects** answer the biological question,
- the **random effects** stop us from pretending that repeated rows from the same cell are independent.

## Why splines are used for time

The original notebook already recognized that many behavioral features change non-linearly across time. A spline is a smooth, flexible way to model that curve without forcing one single straight line.

```{r fig.width=10, fig.height=4.5}
toy_time_df <- data.frame(timepoint = seq(0, 100, by = 1))
spline_basis <- as.data.frame(splines::ns(toy_time_df$timepoint, df = 5))
names(spline_basis) <- paste0("Basis_", seq_len(ncol(spline_basis)))

spline_plot_df <- cbind(toy_time_df, spline_basis) |>
  tidyr::pivot_longer(
    cols = starts_with("Basis_"),
    names_to = "basis",
    values_to = "value"
  )

ggplot(spline_plot_df, aes(timepoint, value, color = basis)) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    title = "Natural spline basis functions for time",
    subtitle = "The model combines these basis functions to represent a curved time trajectory",
    x = "Timepoint",
    y = "Basis value",
    color = "Spline basis"
  )
```

The important interpretation is:

- one spline basis function by itself is not the biological answer,
- but together they create a flexible representation of time,
- and the interaction with `cell_type` asks whether ET bends differently over time than CM.

## How to interpret coefficients without overclaiming

The safest short explanations are:

- `cell_typeET`: ET-vs-CM difference near the reference end of the modeled time axis.
- `cell_typeET:ns(timepoint, 5)k`: part of the ET-vs-CM difference in how the trajectory changes shape over time.

Two cautions matter in a review:

- For spline terms, a single coefficient rarely tells the whole biological story in isolation.
- The most faithful interpretation is often the **overall fitted curve**, not one spline coefficient by itself.

## Why each feature is standardized before fitting

These behavioral features live on very different scales:

- distances can be large,
- ratios are bounded between 0 and 1,
- angles are measured in degrees.

Standardizing each feature before fitting helps because:

- coefficients become more comparable across features,
- model fitting is numerically more stable,
- and the heatmap can use one common color scale more sensibly.

## Model output

```{r}
scrollable_kable(
  analysis_state$model_results,
  caption = "Mixed-model results",
  digits = 4
)
```

```{r fig.width=4, fig.asp=5}
model_heatmap <- analysis_state$model_results
model_heatmap$Param2 <- paste(model_heatmap$Param, model_heatmap$Effect, sep = ".")
rownames(model_heatmap) <- model_heatmap$Param2

model_heatmap$signif_code <- ifelse(
  model_heatmap$pval.adj < 0.001, "***",
  ifelse(model_heatmap$pval.adj < 0.01, "**", ifelse(model_heatmap$pval.adj < 0.05, "*", " "))
)

heatmap_table <- model_heatmap |>
  dplyr::select(Param, Effect, Coef, signif_code) |>
  tidyr::pivot_wider(
    names_from = Effect,
    values_from = c(Coef, signif_code),
    names_sep = "."
  )

coef_columns <- grep("^Coef\\.", names(heatmap_table), value = TRUE)
coef_matrix <- as.matrix(heatmap_table[, coef_columns, drop = FALSE])
rownames(coef_matrix) <- heatmap_table$Param

colour_function <- circlize::colorRamp2(c(-2, 0, 2), c("blue", "white", "red"))

draw(
  Heatmap(
    coef_matrix,
    name = "Coefficient",
    col = colour_function,
    row_names_gp = grid::gpar(fontsize = 8),
    column_names_gp = grid::gpar(fontsize = 10),
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid::grid.text(
        heatmap_table[
          i,
          gsub(
            "^Coef\\.",
            "signif_code.",
            coef_columns[j]
          )
        ],
        x = x,
        y = y,
        gp = grid::gpar(fontsize = 10)
      )
    }
  )
)
```

How to talk through the heatmap:

- The color encodes the coefficient sign and magnitude on the standardized scale.
- Stars encode the Holm-adjusted significance level.
- Rows are parameters.
- Columns are the tested ET main effect and ET-by-spline interaction terms.

## How to answer likely lab-review questions

If someone asks "Why not use a simple linear model?", the answer is:

- because the same cell contributes repeated rows, so the independence assumption would be violated.

If someone asks "Why not fit separate tests at each timepoint?", the answer is:

- because that would throw away the longitudinal structure and make the results noisier and harder to summarize.

If someone asks "What does the random effect buy us biologically?", the answer is:

- it acknowledges that cells do not all start at the same level or change at exactly the same rate, so cell-to-cell heterogeneity does not distort the group comparison.

If someone asks "Why splines?", the answer is:

- because the trajectories are curved over time, and splines let the model capture that curvature smoothly without overfragmenting the analysis.

# Dimensionality Reduction and Clustering

## Color palettes and object construction

```{r}
colors_samples <- scales::hue_pal()(2)
colors_clusters <- c(
  RColorBrewer::brewer.pal(12, name = "Paired"),
  pals::glasbey(32)[-4],
  RColorBrewer::brewer.pal(9, name = "Set1"),
  RColorBrewer::brewer.pal(8, name = "Set2"),
  RColorBrewer::brewer.pal(12, name = "Set3"),
  RColorBrewer::brewer.pal(9, name = "Pastel1"),
  RColorBrewer::brewer.pal(8, name = "Pastel2"),
  RColorBrewer::brewer.pal(8, name = "Accent"),
  RColorBrewer::brewer.pal(8, name = "Dark2")
)
color_expression <- pals::jet(n = 50)

clustering_table <- analysis_state$collapsed_duplicates$clustering_table
categorical_columns <- names(clustering_table)[1:4]
feature_columns <- names(clustering_table)[5:ncol(clustering_table)]

seurat_meta <- clustering_table[, categorical_columns, drop = FALSE]
seurat_meta$node_id <- as.numeric(as.character(seurat_meta$node_id))
seurat_meta$timepoint <- as.numeric(as.character(seurat_meta$timepoint))
seurat_meta <- merge(seurat_meta, analysis_state$duplicate_resolution$corrected_tracks, all.x = TRUE, all.y = FALSE)

tracking_matrix <- t(as.matrix(clustering_table[, feature_columns, drop = FALSE]))
colnames(tracking_matrix) <- rownames(clustering_table)

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_pca.rds"),
  function() {
    object <- suppressWarnings(
      CreateSeuratObject(
        counts = tracking_matrix,
        assay = "IM",
        meta.data = seurat_meta,
        project = "Tracking"
      )
    )
    DefaultAssay(object) <- "IM"
    VariableFeatures(object) <- rownames(object[["IM"]])
    object |>
      ScaleData(verbose = FALSE) |>
      RunPCA(verbose = FALSE)
  }
)
```

Review note:

- Seurat is being used here as a convenient container for PCA, neighbor graph construction, clustering, and 2D embeddings on engineered behavioral features.
- This is not single-cell transcriptomics data, but the abstraction still works for a feature-by-cell matrix.

## PCA and elbow selection

```{r}
std_pc <- Stdev(SO, reduction = "pca")
elbow_point <- estimate_elbow_point(std_pc)
pcas <- seq_len(elbow_point)

ggplot(data.frame(PC = seq_along(std_pc), STD = std_pc), aes(x = PC, y = STD)) +
  geom_point() +
  geom_vline(xintercept = elbow_point, colour = "red") +
  ggtitle(paste("Elbow point:", elbow_point)) +
  theme_bw()
```

How to describe this:

- PCA compresses correlated behavioral features into orthogonal axes.
- The elbow point is used as a practical cutoff for how many PCs to carry forward.
- This preserves the original workflow's intent of automated elbow detection.

## Clustering

```{r}
use_assay <- "IM"
resolutions <- c(0.1, 0.2, 0.5, 0.75)
clustering_names <- paste0(use_assay, "_snn_res.", resolutions)

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_clusters.rds"),
  function() {
    clustered <- FindNeighbors(SO, dims = pcas, reduction = "pca")
    clustered <- FindClusters(clustered, resolution = resolutions, verbose = FALSE)

    for (cluster_name in clustering_names) {
      clustered[[cluster_name]] <- as.factor(
        paste0("C", stringr::str_pad(clustered[[cluster_name]][, 1], 2, pad = "0"))
      )
    }

    clustered
  }
)

cluster_counts <- SO@meta.data[, clustering_names, drop = FALSE] |>
  tidyr::pivot_longer(cols = dplyr::everything(), names_to = "resolution", values_to = "cluster") |>
  dplyr::count(resolution, cluster, name = "n_cells")

knitr::kable(cluster_counts, caption = "Cluster counts across tested resolutions")
```

Why multiple resolutions are shown:

- Louvain clustering depends on the resolution parameter.
- The idea is to inspect coarse and fine partitions before choosing a final grouping for interpretation.
- The refactor keeps the same tested grid as the original notebook.

## tSNE and UMAP parameter grids

```{r}
selected_resolution <- 0.5
resolution_name <- paste0(use_assay, "_snn_res.", selected_resolution)
SO <- SetIdent(SO, value = resolution_name)
DefaultAssay(SO) <- use_assay

selected_perplexity <- 200
selected_neighbors <- 50
selected_min_distance <- 0.3

SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_embeddings_final.rds"),
  function() {
    object <- SO
    object <- RunTSNE(
      object,
      assay = use_assay,
      dims = pcas,
      seed.use = 123456,
      perplexity = selected_perplexity,
      check_duplicates = FALSE,
      reduction.name = "tSNE.Final",
      reduction.key = "tSNE_",
      verbose = FALSE
    )
    object <- RunUMAP(
      object,
      dims = pcas,
      seed.use = 123456,
      n.neighbors = selected_neighbors,
      min.dist = selected_min_distance,
      reduction.name = "UMAP.Final",
      reduction.key = "UMAP_",
      verbose = FALSE
    )
    object
  }
)
```

What to say in the meeting:

- tSNE and UMAP are used as visualization tools, not as the clustering algorithm itself.
- The original workflow explored parameter grids; the refactored notebook directly computes the final selected settings to keep runtime practical on this machine.
- The chosen settings remain the same as the legacy workflow: tSNE perplexity `200`, UMAP neighbors `50`, and UMAP min distance `0.3`.

## Final annotated object and embeddings

```{r}
SO <- load_or_create_rds(
  file.path(paths$cache_dir, "seurat_final.rds"),
  function() {
    final_object <- SetIdent(SO, value = resolution_name)
    final_object$res.Final <- final_object[[resolution_name]][, 1]
    SetIdent(final_object, value = "res.Final")
  }
)

SO <- add_annotations_to_seurat(SO, analysis_state$node_mapping, ntimepoints = 5)
```

One important refactor detail:

- The original notebook tried to recover `cell_type_detailed` from an Excel file on another Windows path.
- The refactor now reads the local replacement directly from `../data/KM1_tracking.xlsx`.
- If that workbook is ever missing, the helper still has the old `track_name`-based fallback so the notebook remains runnable.

```{r fig.width=12, fig.asp=0.35}
DimPlot(
  SO,
  reduction = "tSNE.Final",
  group.by = "cell_type",
  cols = colors_samples,
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "tSNE.Final",
    group.by = "res.Final",
    cols = colors_clusters,
    shuffle = TRUE
  )
```

```{r fig.width=12, fig.asp=0.35}
DimPlot(
  SO,
  reduction = "UMAP.Final",
  group.by = "cell_type",
  cols = colors_samples,
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "UMAP.Final",
    group.by = "res.Final",
    cols = colors_clusters,
    shuffle = TRUE
  )
```

```{r fig.width=12, fig.asp=0.4}
FeaturePlot(
  SO,
  reduction = "UMAP.Final",
  features = c("node_id2", "ntimepoint"),
  combine = TRUE
) &
  scale_colour_gradientn(colors = color_expression, na.value = "grey90")
```

```{r fig.width=12, fig.asp=0.4}
selected_features <- c(
  "trackLength_win1",
  "maxDisplacement_win20",
  "straightness_win20",
  "asphericity_win20",
  "DistBorderMean_win20",
  "overallAngle_win20"
)
selected_features <- gsub("_", "-", selected_features, fixed = TRUE)

FeaturePlot(
  SO,
  reduction = "UMAP.Final",
  features = selected_features[selected_features %in% VariableFeatures(SO)],
  combine = TRUE
) &
  scale_colour_gradientn(colors = color_expression, na.value = "grey90")
```

```{r fig.width=12, fig.asp=0.6}
DimPlot(
  SO,
  reduction = "UMAP.Final",
  group.by = "cell_type_detailed",
  cols = RColorBrewer::brewer.pal(8, "Set1"),
  shuffle = TRUE
) +
  DimPlot(
    SO,
    reduction = "UMAP.Final",
    group.by = "Embryo",
    cols = RColorBrewer::brewer.pal(8, "Set2"),
    shuffle = TRUE
  )
```

# Comparison to Legacy Outputs

```{r}
reference_checks <- dplyr::bind_rows(
  cbind(
    artifact = "clean_wide_nodups",
    compare_with_reference(
      analysis_state$collapsed_duplicates$clustering_table,
      paths$legacy_clean_wide
    )
  ),
  cbind(
    artifact = "longitudinal_stats",
    compare_with_reference(
      analysis_state$model_results,
      paths$legacy_stats
    )
  )
)

knitr::kable(reference_checks, caption = "Structural comparison against legacy reference files")
```

How to discuss this section:

- This comparison is intentionally structural, not byte-for-byte.
- It checks whether the regenerated outputs have the same dimensions and column definitions as the saved legacy artifacts that already exist in `data/`.
- If a future deeper validation is needed, the cached CSVs in `output/behaviors_v16` are the right place to do row-wise numeric comparisons.

# Generated Files and Session Info

```{r}
generated_files <- data.frame(
  file = sort(list.files(paths$output_root, recursive = TRUE))
)

knitr::kable(generated_files, caption = "Files generated by this analysis")
```

```{r}
sessionInfo()
```
