---
title: "C3 Marker Browser Montages for Selected ESC Transition Studies"
subtitle: "Real gene models, BigWig-derived normalization, top up/down C3 peaks, average and replicate-level browser views"
output:
  html_notebook:
    toc: true
    toc_depth: 3
editor_options:
  chunk_output_type: inline
---

This notebook rebuilds the selected-study C3 browser figures using the BigWigs directly.

Key design choices:

1. Peak ranking is computed directly from normalized BigWig signal over strict `C3` marker peaks.
2. Loci are selected programmatically as top `up` and `down` `C3` peaks for each study comparison.
3. Real gene models are drawn with `Gviz::GeneRegionTrack` from `TxDb.Mmusculus.UCSC.mm10.knownGene`.
4. Main figure uses replicate averages for readability.
5. Companion plots keep all replicates.
6. Two scaling modes are saved:
   - `absolute_normalized`: library-size normalized signal, with panel y-limits set from the non-epiblast tracks to avoid flattening everything else
   - `autoscaled`: each track divided by its own 99th percentile, for shape comparison only

## 1. Setup

```{r setup, message=FALSE, warning=FALSE}
if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
renv::activate()

cran_pkgs <- c("tidyverse", "patchwork", "glue", "scales", "png")
bioc_pkgs <- c(
  "GenomicRanges", "rtracklayer", "Signac", "Gviz",
  "TxDb.Mmusculus.UCSC.mm10.knownGene", "org.Mm.eg.db"
)

for (pkg in cran_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
for (pkg in bioc_pkgs) {
  if (!requireNamespace(pkg, quietly = TRUE)) BiocManager::install(pkg, ask = FALSE, update = FALSE)
}

suppressMessages({
  library(tidyverse)
  library(patchwork)
  library(glue)
  library(scales)
  library(png)
  library(grid)
  library(GenomicRanges)
  library(rtracklayer)
  library(Signac)
  library(Gviz)
  library(TxDb.Mmusculus.UCSC.mm10.knownGene)
  library(org.Mm.eg.db)
})

set.seed(42)
options(ucscChromosomeNames = FALSE)
```

## 2. Configuration

```{r config}
DATA_DIR <- file.path("..", "data")
RESULTS_DIR <- file.path("..", "results")
METADATA_FILE <- file.path("..", "ATAC_bigwig_metadata_AIannotated.csv")
MARKER_FILE <- file.path(RESULTS_DIR, "Epiblast_clusters_marker_peaks_annotated_reg.tsv")

OUT_ROOT <- file.path(RESULTS_DIR, "C3_marker_browser_montage_selected_studies")
OUT_TABLES <- file.path(OUT_ROOT, "tables")
OUT_LOCI <- file.path(OUT_ROOT, "locus_panels")
OUT_FIGS <- file.path(OUT_ROOT, "figures")

for (d in c(OUT_ROOT, OUT_TABLES, OUT_LOCI, OUT_FIGS)) {
  if (!dir.exists(d)) dir.create(d, recursive = TRUE)
}

txdb <- TxDb.Mmusculus.UCSC.mm10.knownGene

state_colors <- c("ESC" = "#4D4D4D", "EpiLC" = "#D55E00", "EpiSC" = "#007C91")
study_track_colors <- c(
  "E6.5 Epiblast_C3" = "#6F6258",
  "ESC" = state_colors["ESC"],
  "EpiLC" = state_colors["EpiLC"],
  "EpiSC" = state_colors["EpiSC"],
  "sEpiLC" = "#7B8CCB",
  "d2 EpiLC" = "#F39B6D",
  "EpiLC+CHIR" = "#66A61E",
  "FTW" = "#D95F9A"
)

marker_log2fc_cutoff <- 0.5
marker_padj_cutoff <- 0.05
top_peaks_per_direction <- 3
all_c3_chunk_size <- 6
region_flank_factor <- 10
region_min_flank_bp <- 6000
bin_size_bp <- 25
peak_highlight_fill <- "#F4D03F35"
```

## 3. Helper Functions

```{r helpers}
clean_str <- function(x) {
  x <- ifelse(is.na(x), "", x)
  gsub("\\s+", " ", trimws(x))
}

sanitize_numeric_matrix <- function(mat) {
  mat <- as.matrix(mat)
  storage.mode(mat) <- "double"
  mat[!is.finite(mat)] <- NA_real_
  mat
}

compute_size_factors <- function(mat) {
  mat <- sanitize_numeric_matrix(mat)
  sample_medians <- apply(mat, 1, function(x) median(x[is.finite(x)], na.rm = TRUE))
  pos <- sample_medians[is.finite(sample_medians) & sample_medians > 0]
  gm <- median(pos, na.rm = TRUE)
  if (!is.finite(gm) || gm <= 0) gm <- 1
  sf <- sample_medians / gm
  sf[!is.finite(sf) | sf <= 0] <- 1
  names(sf) <- rownames(mat)
  sf
}

extract_primary_gene <- function(x) {
  xs <- strsplit(ifelse(is.na(x), "", x), "[,;|/ ]+")
  vapply(xs, function(y) {
    y <- y[nzchar(y)]
    if (length(y) == 0) NA_character_ else y[1]
  }, character(1))
}

bw_mean_over_ranges <- function(bw_path, gr) {
  cov <- rtracklayer::import(bw_path, format = "BigWig", which = gr, as = "RleList")
  vals <- rep(NA_real_, length(gr))
  gr_chr <- as.character(seqnames(gr))
  cov_chr <- names(cov)

  mapped <- dplyr::case_when(
    gr_chr %in% cov_chr ~ gr_chr,
    paste0("chr", gr_chr) %in% cov_chr ~ paste0("chr", gr_chr),
    sub("^chr", "", gr_chr) %in% cov_chr ~ sub("^chr", "", gr_chr),
    TRUE ~ NA_character_
  )

  valid_idx <- which(!is.na(mapped))
  if (length(valid_idx) > 0) {
    by_chr <- split(valid_idx, mapped[valid_idx])
    for (chr in names(by_chr)) {
      idx <- by_chr[[chr]]
      v <- Views(cov[[chr]], start = start(gr[idx]), end = end(gr[idx]))
      vals[idx] <- viewMeans(v, na.rm = TRUE)
    }
  }

  vals[is.na(vals)] <- 0
  vals
}

extract_region_signal <- function(bw_path, seq_chr, start_bp, end_bp) {
  gr <- GRanges(seqnames = seq_chr, ranges = IRanges(start = start_bp, end = end_bp))
  cov <- rtracklayer::import(bw_path, format = "BigWig", which = gr, as = "RleList")
  cov_chr <- names(cov)
  mapped_chr <- dplyr::case_when(
    seq_chr %in% cov_chr ~ seq_chr,
    paste0("chr", seq_chr) %in% cov_chr ~ paste0("chr", seq_chr),
    sub("^chr", "", seq_chr) %in% cov_chr ~ sub("^chr", "", seq_chr),
    TRUE ~ NA_character_
  )
  if (is.na(mapped_chr)) return(rep(NA_real_, end_bp - start_bp + 1))
  as.numeric(cov[[mapped_chr]][start_bp:end_bp])
}

bin_signal <- function(signal_vec, chr, from_bp, bin_size = 25) {
  idx <- seq_along(signal_vec)
  bins <- split(idx, ceiling(idx / bin_size))
  tibble(
    seqnames = chr,
    start = from_bp + vapply(bins, min, integer(1)) - 1L,
    end = from_bp + vapply(bins, max, integer(1)) - 1L,
    signal = vapply(bins, function(ii) mean(signal_vec[ii], na.rm = TRUE), numeric(1))
  )
}

make_image_plot <- function(img_path) {
  patchwork::wrap_elements(
    full = grid::rasterGrob(png::readPNG(img_path), interpolate = TRUE)
  )
}

distinct_primary_gene <- function(df) {
  df %>%
    mutate(primary_gene = ifelse(is.na(primary_gene) | primary_gene == "", region, primary_gene)) %>%
    group_by(primary_gene) %>%
    slice_head(n = 1) %>%
    ungroup()
}

render_locus_gviz <- function(
  peak_row,
  track_specs,
  scaling_mode = c("absolute_normalized", "autoscaled"),
  out_png,
  title_prefix,
  compact = FALSE
) {
  scaling_mode <- match.arg(scaling_mode)

  peak_gr <- Signac::StringToGRanges(peak_row$region, sep = c("-", "-"))
  chr <- as.character(seqnames(peak_gr))[1]
  pk_start <- start(peak_gr)[1]
  pk_end <- end(peak_gr)[1]
  pk_width <- width(peak_gr)[1]
  flank <- max(region_min_flank_bp, region_flank_factor * pk_width)
  from_bp <- max(1, pk_start - flank)
  to_bp <- pk_end + flank

  sig_list <- list()
  for (i in seq_len(nrow(track_specs))) {
    bw_vecs <- lapply(seq_along(track_specs$bw_paths[[i]]), function(j) {
      sid <- track_specs$sample_id[[i]][j]
      sf <- track_specs$size_factor[[i]][j]
      raw <- extract_region_signal(track_specs$bw_paths[[i]][j], chr, from_bp, to_bp)
      raw / sf
    })
    sig_mat <- do.call(rbind, bw_vecs)
    if (track_specs$aggregate_mode[i] == "mean") {
      sig <- colMeans(sig_mat, na.rm = TRUE)
    } else {
      sig <- sig_mat[1, ]
    }
    sig_list[[track_specs$track_name[i]]] <- sig
  }

  if (scaling_mode == "autoscaled") {
    sig_list <- lapply(sig_list, function(x) {
      qv <- quantile(x, probs = 0.99, na.rm = TRUE)
      if (!is.finite(qv) || qv <= 0) qv <- max(x, na.rm = TRUE)
      if (!is.finite(qv) || qv <= 0) qv <- 1
      x / qv
    })
    ylim_top <- 1.05
  } else {
    non_epi_names <- setdiff(names(sig_list), "E6.5 Epiblast_C3")
    ref_vals <- unlist(sig_list[non_epi_names], use.names = FALSE)
    if (length(ref_vals) == 0) ref_vals <- unlist(sig_list, use.names = FALSE)
    ylim_top <- quantile(ref_vals, probs = 0.995, na.rm = TRUE)
    if (!is.finite(ylim_top) || ylim_top <= 0) {
      ylim_top <- max(ref_vals, na.rm = TRUE)
    }
    if (!is.finite(ylim_top) || ylim_top <= 0) ylim_top <- 1
  }

  gene_track <- GeneRegionTrack(
    txdb,
    chromosome = chr,
    start = from_bp,
    end = to_bp,
    transcriptAnnotation = "symbol",
    collapseTranscripts = "meta",
    background.title = "white",
    name = "Genes"
  )

  axis_track <- GenomeAxisTrack(name = "")

  peak_track <- AnnotationTrack(
    start = pk_start,
    end = pk_end,
    chromosome = chr,
    genome = "mm10",
    name = "Peak",
    fill = "#D95F02",
    col = "#D95F02"
  )

  displayPars(gene_track) <- list(
    background.title = "white",
    background.panel = "white",
    col.title = "black",
    col.axis = "black",
    cex.title = 0.9,
    fontsize = 11,
    fill = "#24478f",
    col = "#24478f",
    just.group = "above"
  )
  displayPars(axis_track) <- list(background.title = "white", background.panel = "white", cex = 0.7)
  displayPars(peak_track) <- list(background.title = "white", background.panel = "white", cex.title = 0.8)

  data_tracks <- list()
  for (nm in names(sig_list)) {
    this_color <- track_specs$track_color[match(nm, track_specs$track_name)]
    if (length(this_color) == 0 || is.na(this_color)) this_color <- "#4D4D4D"
    binned <- bin_signal(sig_list[[nm]], chr, from_bp, bin_size = bin_size_bp)
    gr <- GRanges(seqnames = binned$seqnames, ranges = IRanges(start = binned$start, end = binned$end))
    tr <- DataTrack(
      range = gr,
      data = matrix(binned$signal, nrow = 1),
      genome = "mm10",
      chromosome = chr,
      type = "histogram",
      baseline = 0,
      name = nm
    )
    displayPars(tr) <- list(
      ylim = c(0, ylim_top),
      background.title = "white",
      background.panel = "white",
      col.axis = "black",
      col.title = "black",
      cex.axis = 0.75,
      cex.title = 0.82,
      lwd = 0.8,
      fill = this_color,
      col.histogram = this_color,
      col = this_color
    )
    data_tracks[[length(data_tracks) + 1]] <- tr
  }

  hl <- HighlightTrack(
    trackList = c(list(axis_track, gene_track, peak_track), data_tracks),
    start = pk_start,
    end = pk_end,
    chromosome = chr,
    fill = peak_highlight_fill,
    col = NA,
    inBackground = TRUE
  )

  if (compact) {
    track_sizes <- c(0.22, 0.55, 0.20, rep(0.68, length(data_tracks)))
    plot_height <- max(620, 170 + 105 * length(data_tracks))
  } else {
    track_sizes <- c(0.35, 0.90, 0.35, rep(1.0, length(data_tracks)))
    plot_height <- max(900, 260 + 180 * length(data_tracks))
  }

  png(out_png, width = 1800, height = plot_height, res = 200)
  plotTracks(
    hl,
    from = from_bp,
    to = to_bp,
    sizes = track_sizes,
    transcriptAnnotation = "symbol",
    background.panel = "white",
    background.title = "white",
    col.axis = "black",
    cex.title = ifelse(compact, 0.68, 0.8),
    main = glue("{title_prefix} | {peak_row$primary_gene} | {peak_row$region}")
  )
  dev.off()
}

make_track_specs <- function(study_meta, epiblast_meta, def, size_factor_tbl, mode = c("average", "replicates")) {
  mode <- match.arg(mode)

  if (mode == "average") {
    out <- list(
      tibble(
        track_name = "E6.5 Epiblast_C3",
        track_color = unname(study_track_colors["E6.5 Epiblast_C3"]),
        bw_paths = list(epiblast_meta$bw_path),
        sample_id = list(epiblast_meta$sample_id),
        size_factor = list(size_factor_tbl$size_factor[match(epiblast_meta$sample_id, size_factor_tbl$sample_id)]),
        aggregate_mode = "mean"
      )
    )

    for (grp in def$group_order) {
      samp <- study_meta %>% filter(sample_group == grp)
      out[[length(out) + 1]] <- tibble(
        track_name = grp,
        track_color = unname(study_track_colors[grp]),
        bw_paths = list(samp$bw_path),
        sample_id = list(samp$sample_id),
        size_factor = list(size_factor_tbl$size_factor[match(samp$sample_id, size_factor_tbl$sample_id)]),
        aggregate_mode = "mean"
      )
    }
    bind_rows(out)
  } else {
    epi_row <- tibble(
      track_name = "E6.5 Epiblast_C3",
      track_color = unname(study_track_colors["E6.5 Epiblast_C3"]),
      bw_paths = list(epiblast_meta$bw_path),
      sample_id = list(epiblast_meta$sample_id),
      size_factor = list(size_factor_tbl$size_factor[match(epiblast_meta$sample_id, size_factor_tbl$sample_id)]),
      aggregate_mode = "single"
    )

    rep_rows <- study_meta %>%
      mutate(track_name = sample_title, track_color = unname(study_track_colors[sample_group])) %>%
      transmute(
        track_name,
        track_color,
        bw_paths = purrr::map(bw_path, ~.x),
        sample_id = purrr::map(sample_id, ~.x),
        size_factor = purrr::map(sample_id, ~size_factor_tbl$size_factor[match(.x, size_factor_tbl$sample_id)]),
        aggregate_mode = "single"
      )

    bind_rows(epi_row, rep_rows)
  }
}

assemble_montage <- function(img_paths, title_text, subtitle_text, ncol = 3) {
  if (length(img_paths) == 0) return(NULL)
  patchwork::wrap_plots(lapply(img_paths, make_image_plot), ncol = ncol) +
    plot_annotation(title = title_text, subtitle = subtitle_text)
}
```

## 4. Load Metadata and Marker Peaks

```{r load-data}
meta <- readr::read_delim(METADATA_FILE, delim = ";", show_col_types = FALSE) %>%
  mutate(
    sample_id = ifelse(is.na(basename) | basename == "", sub("[.]bw$", "", filename), basename),
    sample_title = clean_str(sample_title),
    bw_path = file.path(DATA_DIR, "ESC_other", filename),
    bw_exists = file.exists(bw_path)
  )

marker_tbl_all <- read.delim(MARKER_FILE, sep = "\t", stringsAsFactors = FALSE) %>%
  filter(avg_log2FC > marker_log2fc_cutoff, p_val_adj < marker_padj_cutoff) %>%
  mutate(primary_gene = extract_primary_gene(closest_TSS))

marker_tbl_c3 <- marker_tbl_all %>%
  filter(cluster == "C3")

all_marker_gr <- Signac::StringToGRanges(marker_tbl_all$region, sep = c("-", "-"))
c3_marker_gr <- Signac::StringToGRanges(marker_tbl_c3$region, sep = c("-", "-"))

epiblast_meta <- tibble(
  sample_id = "Sendra_C3",
  sample_title = "E6.5 Epiblast_C3",
  bw_path = file.path(DATA_DIR, "Epiblast", "Sendra_C3.bw"),
  bw_exists = file.exists(bw_path)
) %>%
  filter(bw_exists)

study_definitions <- list(
  "PRJNA648362" = list(
    study_title = "PRJNA648362 | ESC -> EpiLC -> EpiSC",
    comparison_a = "EpiSC",
    comparison_b = "ESC",
    group_order = c("ESC", "EpiLC", "EpiSC"),
    sample_map = c(
      "ATAC_ESC_rep1" = "ESC",
      "ATAC_ESC_rep2" = "ESC",
      "ATAC_EpiLC_rep1" = "EpiLC",
      "ATAC_EpiLC_rep2" = "EpiLC",
      "ATAC_EpiSC_rep1" = "EpiSC",
      "ATAC_EpiSC_rep2" = "EpiSC"
    )
  ),
  "PRJNA856446" = list(
    study_title = "PRJNA856446 | sEpiLC -> d2 EpiLC -> EpiLC+CHIR -> FTW",
    comparison_a = "FTW",
    comparison_b = "sEpiLC",
    group_order = c("sEpiLC", "d2 EpiLC", "EpiLC+CHIR", "FTW"),
    sample_map = c(
      "sEpiLC_1" = "sEpiLC",
      "sEpiLC_2" = "sEpiLC",
      "d2 EpiLC_1" = "d2 EpiLC",
      "d2 EpiLC_2" = "d2 EpiLC",
      "EpiLC+CHIR_1" = "EpiLC+CHIR",
      "EpiLC+CHIR_2" = "EpiLC+CHIR",
      "FTW_1" = "FTW",
      "FTW_2" = "FTW"
    )
  )
)

study_meta_list <- lapply(names(study_definitions), function(study_id) {
  def <- study_definitions[[study_id]]
  meta %>%
    filter(study_accession == study_id, bw_exists, sample_title %in% names(def$sample_map)) %>%
    mutate(sample_group = unname(def$sample_map[sample_title])) %>%
    arrange(factor(sample_group, levels = def$group_order), sample_title)
})
names(study_meta_list) <- names(study_definitions)

cat("Strict markers all clusters:", nrow(marker_tbl_all), "\n")
cat("Strict C3 markers:", nrow(marker_tbl_c3), "\n")
```

## 5. Direct BigWig-Derived Normalization

```{r normalization, message=FALSE, warning=FALSE}
all_selected_samples <- bind_rows(study_meta_list) %>%
  distinct(sample_id, .keep_all = TRUE)

sample_signal_rows <- list()
for (i in seq_len(nrow(all_selected_samples))) {
  sid <- all_selected_samples$sample_id[i]
  bw <- all_selected_samples$bw_path[i]
  sample_signal_rows[[sid]] <- bw_mean_over_ranges(bw, all_marker_gr)
}

epiblast_all_marker <- bw_mean_over_ranges(epiblast_meta$bw_path[1], all_marker_gr)

norm_mat <- do.call(rbind, c(sample_signal_rows, list(Sendra_C3 = epiblast_all_marker)))
rownames(norm_mat) <- c(all_selected_samples$sample_id, "Sendra_C3")
colnames(norm_mat) <- marker_tbl_all$region

size_factor <- compute_size_factors(norm_mat)
size_factor_tbl <- tibble(sample_id = names(size_factor), size_factor = unname(size_factor))
readr::write_csv(size_factor_tbl, file.path(OUT_TABLES, "bigwig_direct_size_factors_selected_studies.csv"))
```

## 6. Rank C3 Peaks Per Study

```{r rank-peaks, message=FALSE, warning=FALSE}
c3_signal_rows <- list()
all_rank_rows <- list()
all_selected_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  peak_cols <- marker_tbl_c3$region

  sample_peak_means <- lapply(seq_len(nrow(sm)), function(i) {
    sid <- sm$sample_id[i]
    sf <- size_factor_tbl$size_factor[match(sid, size_factor_tbl$sample_id)]
    vals <- bw_mean_over_ranges(sm$bw_path[i], c3_marker_gr) / sf
    names(vals) <- peak_cols
    tibble(
      sample_id = sid,
      sample_title = sm$sample_title[i],
      sample_group = sm$sample_group[i]
    ) %>%
      bind_cols(as_tibble_row(as.list(vals)))
  })

  sample_peak_df <- bind_rows(sample_peak_means)
  peak_cols <- marker_tbl_c3$region

  mean_by_group <- sample_peak_df %>%
    group_by(sample_group) %>%
    summarise(across(all_of(peak_cols), ~mean(.x, na.rm = TRUE)), .groups = "drop")

  vec_a <- as.numeric(mean_by_group[mean_by_group$sample_group == def$comparison_a, peak_cols, drop = TRUE])
  vec_b <- as.numeric(mean_by_group[mean_by_group$sample_group == def$comparison_b, peak_cols, drop = TRUE])

  rank_tbl <- marker_tbl_c3 %>%
    transmute(
      study_accession = study_id,
      primary_gene,
      region,
      avg_log2FC_marker = avg_log2FC,
      marker_padj = p_val_adj,
      mean_group_a = vec_a,
      mean_group_b = vec_b,
      log2FC_groupA_vs_groupB = log2((vec_a + 1e-3) / (vec_b + 1e-3)),
      delta_groupA_minus_groupB = vec_a - vec_b,
      higher_accessibility_condition = ifelse(log2FC_groupA_vs_groupB >= 0, def$comparison_a, def$comparison_b),
      lower_accessibility_condition = ifelse(log2FC_groupA_vs_groupB >= 0, def$comparison_b, def$comparison_a)
    ) %>%
    arrange(desc(log2FC_groupA_vs_groupB))

  readr::write_csv(rank_tbl, file.path(OUT_TABLES, paste0(study_id, "_ranked_C3_peaks.csv")))
  all_rank_rows[[study_id]] <- rank_tbl

  selected_up <- rank_tbl %>%
    filter(log2FC_groupA_vs_groupB > 0) %>%
    arrange(desc(log2FC_groupA_vs_groupB), desc(mean_group_a)) %>%
    distinct_primary_gene() %>%
    slice_head(n = top_peaks_per_direction) %>%
    mutate(direction_class = "up")

  selected_down <- rank_tbl %>%
    filter(log2FC_groupA_vs_groupB < 0) %>%
    arrange(log2FC_groupA_vs_groupB, desc(mean_group_b)) %>%
    distinct_primary_gene() %>%
    slice_head(n = top_peaks_per_direction) %>%
    mutate(direction_class = "down")

  selected_tbl <- bind_rows(selected_up, selected_down) %>%
    mutate(
      study_title = def$study_title,
      comparison_label = paste0(def$comparison_a, " vs ", def$comparison_b)
    )

  readr::write_csv(selected_tbl, file.path(OUT_TABLES, paste0(study_id, "_selected_top_C3_peaks.csv")))
  all_selected_rows[[study_id]] <- selected_tbl
}

ranked_tbl_all <- bind_rows(all_rank_rows)
selected_tbl_all <- bind_rows(all_selected_rows)

readr::write_csv(ranked_tbl_all, file.path(OUT_TABLES, "all_selected_studies_ranked_C3_peaks.csv"))
readr::write_csv(selected_tbl_all, file.path(OUT_TABLES, "all_selected_studies_top_C3_peaks.csv"))

selected_tbl_all
```

## 7. Render Real Browser Panels Per Locus

```{r render-locus-panels, message=FALSE, warning=FALSE}
render_manifest_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  sel_tbl <- all_selected_rows[[study_id]]

  avg_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "average")
  rep_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "replicates")

  study_dir <- file.path(OUT_LOCI, study_id)
  if (!dir.exists(study_dir)) dir.create(study_dir, recursive = TRUE)

  for (i in seq_len(nrow(sel_tbl))) {
    peak_row <- sel_tbl[i, ]
    gene_tag <- ifelse(is.na(peak_row$primary_gene) | peak_row$primary_gene == "", paste0("peak", i), peak_row$primary_gene)
    dir_tag <- peak_row$direction_class
    idx_tag <- sprintf("%02d", i)

    avg_abs_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_absolute.png"))
    avg_auto_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_average_autoscaled.png"))
    rep_abs_png <- file.path(study_dir, paste0(idx_tag, "_", dir_tag, "_", gene_tag, "_replicates_absolute.png"))

    render_locus_gviz(
      peak_row = peak_row,
      track_specs = avg_specs,
      scaling_mode = "absolute_normalized",
      out_png = avg_abs_png,
      title_prefix = paste0(study_id, " | average")
    )

    render_locus_gviz(
      peak_row = peak_row,
      track_specs = avg_specs,
      scaling_mode = "autoscaled",
      out_png = avg_auto_png,
      title_prefix = paste0(study_id, " | average autoscaled")
    )

    render_locus_gviz(
      peak_row = peak_row,
      track_specs = rep_specs,
      scaling_mode = "absolute_normalized",
      out_png = rep_abs_png,
      title_prefix = paste0(study_id, " | replicates")
    )

    render_manifest_rows[[length(render_manifest_rows) + 1]] <- tibble(
      study_accession = study_id,
      direction_class = peak_row$direction_class,
      primary_gene = peak_row$primary_gene,
      region = peak_row$region,
      average_absolute_png = avg_abs_png,
      average_autoscaled_png = avg_auto_png,
      replicates_absolute_png = rep_abs_png
    )
  }
}

render_manifest <- bind_rows(render_manifest_rows)
readr::write_csv(render_manifest, file.path(OUT_TABLES, "render_manifest_C3_browser_panels.csv"))
render_manifest
```

## 8. Assemble Study Montages

```{r montage-assembly, fig.width=15, fig.height=10, message=FALSE, warning=FALSE}
study_montage_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  man <- render_manifest %>% filter(study_accession == study_id)
  if (nrow(man) == 0) next

  up_avg_abs <- man %>% filter(direction_class == "up") %>% pull(average_absolute_png)
  down_avg_abs <- man %>% filter(direction_class == "down") %>% pull(average_absolute_png)
  up_avg_auto <- man %>% filter(direction_class == "up") %>% pull(average_autoscaled_png)
  down_avg_auto <- man %>% filter(direction_class == "down") %>% pull(average_autoscaled_png)

  abs_plot <- (
    assemble_montage(
      up_avg_abs,
      title_text = paste0("More accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, library-size normalized"
    ) /
    assemble_montage(
      down_avg_abs,
      title_text = paste0("Less accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, library-size normalized"
    )
  ) + plot_annotation(
    title = def$study_title,
    subtitle = "Peak ranking and normalization derived directly from the BigWigs | y-limits set from non-epiblast tracks to avoid epiblast flattening"
  )

  auto_plot <- (
    assemble_montage(
      up_avg_auto,
      title_text = paste0("More accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, per-track autoscaled to the 99th percentile"
    ) /
    assemble_montage(
      down_avg_auto,
      title_text = paste0("Less accessible in ", def$comparison_a, " vs ", def$comparison_b),
      subtitle_text = "Replicate-average tracks, per-track autoscaled to the 99th percentile"
    )
  ) + plot_annotation(
    title = paste0(def$study_title, " | autoscaled"),
    subtitle = "Shape-comparison view only; amplitudes should not be interpreted quantitatively"
  )

  abs_png <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_absolute.png"))
  auto_png <- file.path(OUT_FIGS, paste0(study_id, "_C3_top_up_down_average_autoscaled.png"))

  ggsave(abs_png, abs_plot, width = 15, height = 10, dpi = 450)
  ggsave(auto_png, auto_plot, width = 15, height = 10, dpi = 450)

  study_montage_rows[[length(study_montage_rows) + 1]] <- tibble(
    study_accession = study_id,
    study_title = def$study_title,
    average_absolute_png = abs_png,
    average_autoscaled_png = auto_png
  )
}

study_montage_manifest <- bind_rows(study_montage_rows)
readr::write_csv(study_montage_manifest, file.path(OUT_TABLES, "study_montage_manifest.csv"))
study_montage_manifest
```

## 9. All C3 Peaks, Autoscaled, Grouped in Montages of Six

```{r all-c3-autoscaled-compact, message=FALSE, warning=FALSE}
all_c3_rows <- list()

for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sm <- study_meta_list[[study_id]]
  avg_specs <- make_track_specs(sm, epiblast_meta, def, size_factor_tbl, mode = "average")
  rank_tbl <- all_rank_rows[[study_id]] %>%
    arrange(desc(log2FC_groupA_vs_groupB), desc(abs(delta_groupA_minus_groupB))) %>%
    mutate(order_idx = row_number())

  compact_dir <- file.path(OUT_LOCI, study_id, "all_c3_autoscaled_compact")
  if (!dir.exists(compact_dir)) dir.create(compact_dir, recursive = TRUE)

  compact_paths <- character(nrow(rank_tbl))
  for (i in seq_len(nrow(rank_tbl))) {
    peak_row <- rank_tbl[i, ]
    gene_tag <- ifelse(is.na(peak_row$primary_gene) | peak_row$primary_gene == "", paste0("peak", i), peak_row$primary_gene)
    compact_paths[i] <- out_png
    if (!file.exists(out_png)) {
      render_locus_gviz(
        peak_row = peak_row,
        track_specs = avg_specs,
        scaling_mode = "autoscaled",
        out_png = out_png,
        title_prefix = paste0(study_id, " | compact autoscaled"),
        compact = TRUE
      )
    }
  }

  chunks <- split(compact_paths, ceiling(seq_along(compact_paths) / all_c3_chunk_size))
  for (page_idx in seq_along(chunks)) {
    montage_plot <- assemble_montage(
      chunks[[page_idx]],
      title_text = paste0(def$study_title, " | all strict C3 peaks | autoscaled"),
      subtitle_text = paste0(
        "Page ", page_idx, " of ", length(chunks),
        " | loci ordered by ", def$comparison_a, " vs ", def$comparison_b, " log2FC"
      ),
      ncol = 3
    )
    montage_png <- file.path(OUT_FIGS, paste0(study_id, "_all_C3_autoscaled_page_", sprintf("%02d", page_idx), ".png"))
    montage_svg <- file.path(OUT_FIGS, paste0(study_id, "_all_C3_autoscaled_page_", sprintf("%02d", page_idx), ".svg"))

    if (!file.exists(montage_png)) {
      ggsave(montage_png, montage_plot, width = 15, height = 9, dpi = 350)
    }

    if (!file.exists(montage_svg)) {
      ggsave(montage_svg, montage_plot, width = 15, height = 9)
    }

    all_c3_rows[[length(all_c3_rows) + 1]] <- tibble(
      study_accession = study_id,
      study_title = def$study_title,
      page = page_idx,
      n_pages = length(chunks),
      montage_png = montage_png
    )
  }
}

all_c3_montage_manifest <- bind_rows(all_c3_rows)
readr::write_csv(all_c3_montage_manifest, file.path(OUT_TABLES, "all_C3_autoscaled_compact_montage_manifest.csv"))
all_c3_montage_manifest
```

## 10. All C3 Peak Montages in Report

```{r all-c3-in-report, results='asis', echo=FALSE}
for (study_id in names(study_definitions)) {
  def <- study_definitions[[study_id]]
  sub_man <- all_c3_montage_manifest %>% filter(study_accession == study_id)
  cat("\n\n### ", def$study_title, " | all strict C3 peaks autoscaled\n\n", sep = "")
  for (i in seq_len(nrow(sub_man))) {
    cat("#### Montage page ", sub_man$page[i], " / ", sub_man$n_pages[i], "\n\n", sep = "")
    print(knitr::include_graphics(sub_man$montage_png[i]))
    cat("\n\n")
  }
}
```

## 11. Notes on Interpretation

```{r notes}
cat(
  paste(
    "- Main study figures use replicate averages because that is the clearest way to compare trajectory-level accessibility without overloading the panel.",
    "- Replicate-level browser plots are still saved for every selected locus in the locus_panels directory.",
    "- Real gene models come from Gviz + TxDb.Mmusculus.UCSC.mm10.knownGene.",
    "- Absolute normalized plots preserve between-group amplitude after size-factor normalization from the BigWigs.",
    "- Autoscaled plots divide each displayed track by its own 99th percentile and are meant only for shape comparison.",
    sep = "\n"
  )
)
```

## 12. Outputs

```{r outputs}
cat("Tables:\n")
cat("-", file.path(OUT_TABLES, "bigwig_direct_size_factors_selected_studies.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_ranked_C3_peaks.csv"), "\n")
cat("-", file.path(OUT_TABLES, "all_selected_studies_top_C3_peaks.csv"), "\n")
cat("-", file.path(OUT_TABLES, "render_manifest_C3_browser_panels.csv"), "\n")
cat("-", file.path(OUT_TABLES, "study_montage_manifest.csv"), "\n\n")
cat("-", file.path(OUT_TABLES, "all_C3_autoscaled_compact_montage_manifest.csv"), "\n\n")

cat("Study montages:\n")
cat("-", file.path(OUT_FIGS, "PRJNA648362_C3_top_up_down_average_absolute.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA648362_C3_top_up_down_average_autoscaled.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA856446_C3_top_up_down_average_absolute.png"), "\n")
cat("-", file.path(OUT_FIGS, "PRJNA856446_C3_top_up_down_average_autoscaled.png"), "\n\n")
cat("All C3 compact autoscaled montages:\n")
cat("-", OUT_FIGS, "\n\n")

cat("Per-locus panels:\n")
cat("-", OUT_LOCI, "\n")
```

## 13. Session Info

```{r session-info}
sessionInfo()
```
